---
title: "data_analysis_biosamples_psychopathology"
author: "Nico Bast"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    theme: cerulean
    code_folding: hide
  word_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#define paths
root_path<-'C:/Users/nico/Nextcloud/project_sega' #local data path - masterfile
data_path<-'//192.168.88.212/daten/KJP_Studien/Studie_SEGA' #remote data path - WINET - requires VPN to data
project_path<-"C:/Users/nico/Nextcloud/Paper_SEGA_stress" #project folder


#required packages
require(openxlsx) #read xls
require(ggplot2) #visuals
require(gridExtra)
require(effectsize) #standardized coefficients of LM
require(kableExtra) #APA tables
require(corrplot) #correlation plot
require(car) # ANOVA type=3
require(emmeans) # post hoc
require(caret) #classification
require(pROC) #classification
require(boot) #classification
require(tidyr) #data manipulation
library(dplyr) #data manipulation
library(psych) #model diagnostics
library(GPArotation) #exploratory factor analysis
library(plotly) #make figure zoomable/scalable in VS code

#set ggplot theme
theme_set(theme_bw())

#custom functions

#customaized ANOVA output
analyze_lm<-function(x,log_backtransform=F){
  
  model_summary<-summary(x)
  print(paste('adj. R²:',round(model_summary$adj.r.squared,2)))
  
  print(car::Anova(x,type=3))
  return(cbind(round(summary(x)$coefficients[,1],2),round(confint(x),2)))
  
  if(log_backtransform){
  return(cbind(round(10^summary(x)$coefficients[,1],2),round(10^confint(x),2)))
  }
  
}

#create tables for supplements
save_model<-function(lm_output=lm,
                     var_names=c(attr(lm$coefficients,'names'),'Residuals'),
                     caption_label,
                     figure_name){

  
#table 1 
table_data<-round(car::Anova(lm_output,type=3),3)
table_data<-cbind(var_names,table_data)

table_output<-table_data %>%
  kbl(caption = paste0('Model summary - ',caption_label),
      col.names = c('','Sum Sq','DF','F-value','p-value'),
      row.names = F) %>%
  kable_classic(full_width = F, html_font = "Cambria")

save_kable(table_output, file=paste0(project_path,'/output/supplements/',figure_name,'_1.html'))

#table 2
table_data2<-data.frame(cbind(round(summary(lm_output)$coefficients[,1],2),round(confint(lm_output),2)))
table_data2<-cbind(head(var_names,-1),table_data2)

table_output<-table_data2 %>%
  kbl(caption = paste0('Effect estimates - ',caption_label),
      col.names = c('','estimate (beta)','lower bound (2.5%)','upper bound (97.5%)'),
      row.names = F) %>%
  kable_classic(full_width = F, html_font = "Cambria")

save_kable(table_output, file=paste0(project_path,'/output/supplements/',figure_name,'_2.html'))

}

```

# SEGA biosamples psychopathology - stress

Aim: Relate biosample data (hair and saliva) and pupillometry (baseline pupil size [BPS] and stimulus‑evoked pupillary response [SEPR]) to transdiagnostic psychopathology in an autism‑enriched sample.

## data assembly

- combine different data sources and calculate core variables

### load data

- utilizes a master file of the SEGA data that contains already preprocessed data

```{r read data}

#main data frame as outcome from SEGA data
df<-read.xlsx(paste0(root_path,'/data/master_file_SEGA_Oct2025.xlsx'),na.strings = c("NA","-"),detectDates = T)

#experimental data including SDQ and further cortisol data
df_exp<-read.xlsx(paste0(root_path,'/data/master_file_SEGA_data_Sept2025.xlsx'),startRow = 3)
df_exp<-df_exp[,c(1,3,18:29,42:43,47:56,57:88)] #retain matching, SDQ, height/weight, cortisol, and experimental data

# #comorbidites
df_comorbidity<-read.xlsx(paste0(root_path,'/data/comorbidities_SEGA.xlsx'),detectDates = T)

# baseline pupil size data from raw pupil data
df_baseline<-readRDS(paste0(root_path,'/data/baseline_means.rds'))

```

### remodel baseline data

```{r remodel baseline data}

#remodel baseline data to match the format

df_baseline<-df_baseline[,names(df_baseline)!='block_counter']

df_baseline <- df_baseline %>%
  tidyr::pivot_wider(
    id_cols = id,
    names_from = baseline_trial_counter,
    values_from = block_baseline_mean
  )


#allign SEGA_ID with other dataframes
names(df_baseline)<-c('SEGA_ID',paste0('BPS_block_',1:6))
df_baseline$sega_id <- sprintf("SEGA_%03d", df_baseline$SEGA_ID)

```

### rename variables

```{r rename variables}

#saliva samples renaming
names(df_exp)[17:26]<-paste0('saliva_',names(df_exp)[17:26])

names(df)[42:45]<-c('saliva_Value.1','saliva_Value.2','saliva_CV.1','saliva_CV.2')

#hair renaming
names(df)[38]<-'hair_mass_mg'
names(df)[39]<-'hair_cortisol_pg_mg'
names(df)[40]<-'hair_cortisone_pg_mg'

#SDQ
names(df_exp)[3:8]<-c('SDQ_total','SDQ_emo','SDQ_beh','SDQ_hyp','SDQ_beh_peers','SDQ_pros')
names(df_exp)[9:14]<-c('SDQ_total','SDQ_emo','SDQ_beh','SDQ_hyp','SDQ_beh_peers','SDQ_pros')

names(df_exp)[3:8]<-paste0(names(df_exp)[3:8],'_self')
names(df_exp)[9:14]<-paste0(names(df_exp)[9:14],'_parent')

```

### recode time

- allows to investigate time of day for hormone analysis

```{r recode time}

#the magic number 86400 is the number of seconds in a day. Multiplying the Excel serial number by this gives us seconds since the Excel epoch ("1899-12-30"). as.POSIXct converts these seconds to a datetime object, and format(..., "%H:%M") extracts just the time string.

# #convert to string "HH:MM" for each time point
# df$saliva_Time.1 <- format(as.POSIXct(df$saliva_Time.1 * 86400, origin = "1899-12-30", tz = "UTC"), "%H:%M")
# df$saliva_Time.2 <- format(as.POSIXct(df$saliva_Time.2 * 86400, origin = "1899-12-30", tz = "UTC"), "%H:%M")

# #convert to numeric fraction of day for each time point
# df$saliva_Time.1 <- as.numeric(difftime(
#   time1 = as.POSIXct(df$saliva_Time.1, format = "%H:%M"), 
#   time2 = as.POSIXct("00:00", format = "%H:%M"), 
#   units = "secs"
# )) / 86400

# df$saliva_Time.2 <- as.numeric(difftime(
#   time1 = as.POSIXct(df$saliva_Time.2, format = "%H:%M"), 
#   time2 = as.POSIXct("00:00", format = "%H:%M"), 
#   units = "secs"
# )) / 86400

#convert to string "HH:MM" for each time point
df_exp$saliva_Time.1 <- format(as.POSIXct(df_exp$saliva_Time.1 * 86400, origin = "1899-12-30", tz = "UTC"), "%H:%M")
df_exp$saliva_Time.2 <- format(as.POSIXct(df_exp$saliva_Time.2 * 86400, origin = "1899-12-30", tz = "UTC"), "%H:%M")

#convert to numeric fraction of day for each time point
df_exp$saliva_Time.1 <- as.numeric(difftime(
  time1 = as.POSIXct(df_exp$saliva_Time.1, format = "%H:%M"), 
  time2 = as.POSIXct("00:00", format = "%H:%M"), 
  units = "secs"
)) / 86400

df_exp$saliva_Time.2 <- as.numeric(difftime(
  time1 = as.POSIXct(df_exp$saliva_Time.2, format = "%H:%M"), 
  time2 = as.POSIXct("00:00", format = "%H:%M"), 
  units = "secs"
)) / 86400


```

### merge data

```{r merge}


#merge demographics and experimental data
#df<-merge(df,df_exp,by.x='SEGA_ID',by.y='ID_Studie')
df$sega_id <- sprintf("SEGA_%03s", df$sega_id)
df$sega_id<-gsub(' ','0',df$sega_id)
df<-merge(df,df_exp,by.x='sega_id',by.y='ID_Studie')

#merge CBCL factor scores - 4 participants unmatched
#df<-merge(df,df_CBCL_f,by.x='ID-BADO',by.y='ID_Bado')

#merge df_baseline - 10 unmatched
df<-merge(df,df_baseline,by='sega_id',all.x=T)



```

### calculate age

```{r add age}

#merge age
df$age<-as.numeric(with(df,date_data_collection-birthday)/365)


```

### add comorbidity data

- one or more comorbidities

```{r add comorbidity}

df_comorbidity$sega_id <- sprintf("SEGA_%03s", df_comorbidity$SEGA_ID)
df_comorbidity$sega_id<-gsub(' ','0',df_comorbidity$sega_id)

df_comorbidity<-df_comorbidity[,names(df_comorbidity)!='group']

df<-merge(df,df_comorbidity,by='sega_id')


```

## data manipulation

### create saliva change variable

```{r}

df$saliva_change<-df$saliva_Average.2-df$saliva_Average.1
hist(df$saliva_change)

```

### recode group variable

```{r group_recode}

table(df$group)
df$autism_boolean<-ifelse(df$group=='ASD',T,F)

```

### logarithmize bio samples

```{r log_transform_biosamples}

df$saliva_Average.1_log<-log10(df$saliva_Average.1)
df$saliva_Average.2_log<-log10(df$saliva_Average.2)
df$hair_cortisol_pg_mg_log<-log10(df$hair_cortisol_pg_mg)

df$saliva_change_log<-df$saliva_Average.2_log-df$saliva_Average.1_log
df$saliva_change<-df$saliva_Average.2-df$saliva_Average.1

```

### IQ estimate

- iq estimate based on four Wechsler subtests:
- non-verbal: matrices, block test, 
- verbal: commonalities, passive word knowledge

```{r iq_estimate}

df$iq<-rowMeans(df[,names(df) %in% c('iq_mt','iq_gf','iq_mz','iq_wt')],na.rm=T)
hist(df$iq)

```

### estimate BMI

```{r bmi_estimate}

df$BMI<-with(df,Gewicht/(Größe/100*Größe/100))

```

### pupil measure averages

```{r pupil_averages}

df$BPS_average<-(df$pd_low_oddball_saliencehigh_utilityhigh+
  df$pd_low_oddball_saliencehigh_utilitylow+
  df$pd_low_oddball_saliencelow_utilityhigh+
  df$pd_low_oddball_saliencelow_utilitylow+
  df$BPS_Oddball_before_forward+
  df$BPS_Oddball_after_forward+
  df$BPS_Oddball_before_reverse+
  df$BPS_Oddball_after_reverse)/8


df$SEPR_average<-(df$rpd_oddball_saliencehigh_utilityhigh+
                    df$rpd_oddball_saliencehigh_utilitylow+
                    df$rpd_oddball_saliencelow_utilityhigh+
                    df$rpd_oddball_saliencelow_utilitylow+
                    df$SEPR_Oddball_after_forward+
                    df$SEPR_Oddball_after_reverse+
                    df$SEPR_Oddball_before_forward+
                    df$SEPR_Oddball_before_reverse)/8

hist(df$BPS_average)
hist(df$SEPR_average)  

```

### pupil size changes

```{r pupil_changes}

df$SEPR_Oddball_change<-df$SEPR_Oddball_after_forward-df$SEPR_Oddball_before_forward
df$BPS_change<-df$BPS_Standard_after_forward-df$BPS_Standard_before_forward

```

### pupil variation

```{r pupil_variation}

df$SEPR_variation<- apply(df[, c(
  "SEPR_Oddball_after_forward",
  "SEPR_Oddball_after_reverse",
  "SEPR_Oddball_before_forward",
  "SEPR_Oddball_before_reverse",
  "rpd_oddball_saliencehigh_utilityhigh",
  "rpd_oddball_saliencehigh_utilitylow",
  "rpd_oddball_saliencelow_utilityhigh",
  "rpd_oddball_saliencelow_utilitylow"
)], 1, sd, na.rm = TRUE)

df$BPS_variation <- apply(df[, c(
  "BPS_block_1",
  "BPS_block_2",
  "BPS_block_3",
  "BPS_block_4",
  "BPS_block_5",
  "BPS_block_6",
  "pd_low_oddball_saliencehigh_utilityhigh",
  "pd_low_oddball_saliencehigh_utilitylow",
  "pd_low_oddball_saliencelow_utilityhigh",
  "pd_low_oddball_saliencelow_utilitylow"
)], 1, sd, na.rm = TRUE)

```

### SDQ externalizing internalizing - self report

- use Germans norms
- calculate internalizing and externalizing score

```{r sdq_self_norms}

#norms
norms_emo_self <- data.frame(
  raw = 0:10,
  percentile = c(16, 38, 58, 74, 86, 92, 96, 98, 99, 100, 100)
)

norms_beh_peers_self <- data.frame(
  raw = 0:10,
  percentile = c(16, 43, 68, 84, 93, 97, 99, 100, 100, 100, 100)
)

norms_beh_self <- data.frame(
  raw = 0:10,
  percentile = c(12, 43, 72, 88, 95, 98, 99, 100, 100, 100, 100)
)

norms_hyp_self <- data.frame(
  raw = 0:10,
  percentile = c(6, 16, 31, 49, 66, 83, 91, 97, 99, 100, 100)
)

#translate raw to percentiles based on norms
df$pcts_emo_self  <- norms_emo_self$percentile[match(df$SDQ_emo_self, norms_emo_self$raw) ]
df$pcts_beh_peers_self  <- norms_beh_peers_self$percentile[match(df$SDQ_beh_peers_self, norms_beh_peers_self$raw) ]
df$pcts_beh_self  <- norms_beh_self$percentile[match(df$SDQ_beh_self, norms_beh_self$raw) ]
df$pcts_hyp_self  <- norms_hyp_self$percentile[match(df$SDQ_hyp_self, norms_hyp_self$raw) ]

#PERCENTILES: internalizing
df$SDQ_self_int_pcts<-(df$pcts_emo_self+df$pcts_beh_peers_self)/2
df$SDQ_self_ext_pcts<-(df$pcts_beh_self+df$pcts_hyp_self)/2

#RAW: internalizing 
df$SDQ_self_int<-df$SDQ_emo_self+df$SDQ_beh_peers_self
df$SDQ_self_ext<-df$SDQ_beh_self+df$SDQ_hyp_self

grid.arrange(
  ggplot(df,aes(SDQ_self_int_pcts))+geom_histogram(),
  ggplot(df,aes(SDQ_self_ext_pcts))+geom_histogram(),
  ggplot(df,aes(SDQ_self_int))+geom_histogram(),
  ggplot(df,aes(SDQ_self_ext))+geom_histogram()
)

```

### SDQ externalizing internalizing - parent report

- use Germans norms
- calculate internalizing and externalizing score

```{r sdq_parent_norms}

#norms
norms_emo_parent <- data.frame(
  raw = 0:10,
  percentile = c(16, 38, 58, 74, 86, 92, 96, 98, 99, 100, 100)
)

norms_beh_peers_parent <- data.frame(
  raw = 0:10,
  percentile = c(16, 43, 68, 84, 93, 97, 99, 100, 100, 100, 100)
)

norms_beh_parent <- data.frame(
  raw = 0:10,
  percentile = c(12, 43, 72, 88, 95, 98, 99, 100, 100, 100, 100)
)

norms_hyp_parent <- data.frame(
  raw = 0:10,
  percentile = c(6, 16, 31, 49, 66, 83, 91, 97, 99, 100, 100)
)

#translate raw to percentiles based on norms
df$pcts_emo_parent  <- norms_emo_parent$percentile[match(df$SDQ_emo_parent, norms_emo_parent$raw) ]
df$pcts_beh_peers_parent  <- norms_beh_peers_parent$percentile[match(df$SDQ_beh_peers_parent, norms_beh_peers_parent$raw) ]
df$pcts_beh_parent  <- norms_beh_parent$percentile[match(df$SDQ_beh_parent, norms_beh_parent$raw) ]
df$pcts_hyp_parent  <- norms_hyp_parent$percentile[match(df$SDQ_hyp_parent, norms_hyp_parent$raw) ]

#PERCENTILES: internalizing
df$SDQ_parent_int_pcts<-df$pcts_emo_parent+df$pcts_beh_peers_parent
df$SDQ_parent_ext_pcts<-df$pcts_beh_parent+df$pcts_hyp_parent

#RAW: internalizing 
df$SDQ_parent_int<-df$SDQ_emo_parent+df$SDQ_beh_peers_parent
df$SDQ_parent_ext<-df$SDQ_beh_parent+df$SDQ_hyp_parent

```

### outlier exclusion 

- none are excluded

```{r outlier_exclusion}

#questionnaires
hist(df$YSR_GES)
hist(df$YSR_T_GES)

# #correct typo
# df$YSR_GES[df$YSR_T_GES>100 & !is.na(df$YSR_T_GES)]
# df$YSR_GES[df$YSR_T_GES<25 & !is.na(df$YSR_T_GES)]

# df$YSR_T_GES[which(df$YSR_T_GES>100)]<-48
# df$YSR_T_GES[which(df$YSR_T_GES<25)]<-89

```

## data quality

### missing data

-complete data for CBCL
-more missings in SDQ

```{r missing data}

CBCL<-with(df,table(!is.na(CBCL_T_GES)))
# CBCL_factor<-with(df,table(!is.na(cbcl_P_f)))
# CBCL_factor<-c(FALSE,CBCL_factor)

YSR<-with(df,table(!is.na(YSR_T_GES)))

SDQ_self<-with(df,table(!is.na(SDQ_total_self)))
SDQ_parent<-with(df,table(!is.na(SDQ_total_parent)))

hair_cortisol<-with(df,table(!is.na(hair_cortisol_pg_mg)))
saliva_cortisol_measure1<-with(df,table(!is.na(saliva_Average.1)))
saliva_cortisol_measure2<-with(df,table(!is.na(saliva_Average.2)))

BPS_average_mis<-with(df,table(!is.na(BPS_average)))
SEPR_average_mis<-with(df,table(!is.na(SEPR_average)))

missing_table<-rbind(CBCL, YSR,SDQ_self,SDQ_parent,hair_cortisol,saliva_cortisol_measure1,saliva_cortisol_measure2,
                     BPS_average_mis,SEPR_average_mis)

missing_table %>%
  kbl(caption = "Available data for data modalities") %>%
  kable_classic(full_width = F, html_font = "Cambria")

with(df,table(!is.na(YSR_T_GES),!is.na(hair_cortisol_pg_mg)))
with(df,table(!is.na(SDQ_total_parent),!is.na(hair_cortisol_pg_mg)))
with(df,table(!is.na(hair_cortisol_pg_mg),!is.na(BPS_average)))
with(df,table(!is.na(hair_cortisol_pg_mg),!is.na(saliva_Average.1)))

```

### participants by diagnosis and available biosamples

```{r participants_by_diagnosis}

table(df$group[!is.na(df$hair_cortisol_pg_mg) & !is.na(df$saliva_Average.1)])

```

### bio samples data quality

#### hair cortisol

```{r hair}

ggplot(df,aes(x=hair_cortisone_pg_mg,y=hair_cortisol_pg_mg,color=hair_mass_mg))+
  geom_point()+
  labs(x='hair cortisone (pg/mg, log-scaled)', y='hair cortisol (pg/mg, log-scaled)', color='sample mass (mg)')+
  geom_smooth(color='grey',method='lm')+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

```

#### saliva cortisol

```{r saliva}

hist(df$saliva_Average.1)
hist(df$saliva_Average.2)

ggplot(df,aes(y=saliva_Average.1,x=saliva_CV.1))+
  geom_point()+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

ggplot(df,aes(y=saliva_Average.2,x=saliva_CV.2))+
  geom_point()+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

ggplot(df,aes(x=saliva_Average.2-saliva_Average.1))+
  geom_histogram()+
  geom_vline(xintercept=0,linetype=2)+
  labs(title='saliva cortisol reduction within experiment (pg/mg)')

ggplot(df,aes(x=saliva_Average.1,y=saliva_Average.2,color=saliva_CV.1))+
  geom_point()+
  geom_smooth(color='grey',method='lm')+
  #labs(x='hair cortisone (pg/mg, log-scaled)', y='hair cortisol (pg/mg, log-scaled)', color='sample mass (mg)')+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

```

##### intra- inter-assay coefficient of variation (CV)

```{r intra-assay CV}

psych::describe(df$saliva_CV.1)
psych::describe(df$saliva_CV.2)

```

### bio samples

- suggests to log-transform variables in analysis

```{r data inspection bio samples}

ggplot(df,aes(hair_cortisol_pg_mg))+geom_histogram()
ggplot(df,aes(saliva_Average.1))+geom_histogram()
ggplot(df,aes(saliva_Average.2))+geom_histogram()
ggplot(df,aes(saliva_change))+geom_histogram()

ggplot(df,aes(x=hair_cortisol_pg_mg,y=saliva_change))+geom_point()+
  scale_x_continuous(trans='log10')+
  scale_y_continuous(trans='log10')+
  geom_smooth(method='lm')

```

## data analysis

### sample definition

- define sample as participants with self-reported psychopathology and saliva cortisol

```{r sample definition}

df_sample<-df[!is.na(df$YSR_T_GES) & !is.na(df$saliva_Average.1),]

#total sample
nrow(df_sample)

```

### sample description

```{r sample description}

table(df_sample$gender)

table_data<-rbind(psych::describe(df_sample$age),
                  psych::describe(df_sample$iq),
                  psych::describe(df_sample$YSR_T_GES),
                  psych::describe(df_sample$YSR_T_INT),
                  psych::describe(df_sample$YSR_T_EXT),
                  psych::describe(df_sample$CBCL_T_GES),
                  psych::describe(df_sample$CBCL_T_INT),
                  psych::describe(df_sample$CBCL_T_EXT),
                  psych::describe(df_sample$SDQ_total_self),
                  psych::describe(df_sample$SDQ_self_int),
                  psych::describe(df_sample$SDQ_self_ext),
                  psych::describe(df_sample$SDQ_total_parent),
                  psych::describe(df_sample$SDQ_parent_int),
                  psych::describe(df_sample$SDQ_parent_ext),
                  psych::describe(df_sample$hair_cortisol_pg_mg),
                  psych::describe(df_sample$saliva_Average.1),
                  psych::describe(df_sample$saliva_Average.2))


table_data<-table_data[,c('n','mean','sd','min','max')]
table_data<-round(table_data,2)

row.names(table_data)<-c('age','iq','YSR TOTAL (t)','YSR INT (t)','YSR EXT (t)',
                         'CBCL TOTAL (t)','CBCL INT (t)','CBCL EXT (t)',
                         'SDQ TOTAL self (sum)','SDQ INT self (sum)','SDQ EXT self (sum)',
                         'SDQ TOTAL parent (sum)','SDQ INT parent (sum)','SDQ EXT parent (sum)',
                         'hair cortisol (pg/mg)','saliva cortisol before (nmol/l)','saliva cortisol after (nmol/l)')


table_sample<-table_data %>%
  kbl(caption = "Table 1. Sample description") %>%
  kable_classic(full_width = F, html_font = "Cambria")
  

table_sample

save_kable(table_sample, file=paste0(project_path,'/output/tables/table1_sampledescription.html'))

```

### comorbidities

```{r comorbidites}

#with primary diagnosis
table(df_sample$group=='CON')

#primnary diagnosis
table(df_sample$primary.diagnosis)

#comorbdiity
table(!is.na(df_sample$other.diagnoses))
table(c(df_sample$other.diagnoses,df_sample$other.diagnoses.1))

```


### figure questionnaire data

```{r data inspection questionnaires}

bin_count<-15

tiff(file=paste0(project_path,"/output/figures/figure_questionnaire_distribution.tiff"), 
     width=8, height=4, units="in", res=600, compression='lzw') 

#core questionnaire measures
grid.arrange(
  ggplot(df_sample,aes(YSR_GES,group=Gruppe))+
    geom_histogram(bins=bin_count)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('YSR Total'),
  ggplot(df_sample,aes(YSR_INT,group=Gruppe))+
    geom_histogram(bins=bin_count)+ylim(0,20)+xlim(0,50)+theme(axis.title.y = element_blank())+xlab('YSR Internalizing'),
  ggplot(df_sample,aes(YSR_EXT,group=Gruppe))+
    geom_histogram(bins=bin_count)+ylim(0,32)+xlim(0,50)+theme(axis.title.y = element_blank())+xlab('YSR Externalizing'),
  
  ggplot(df_sample,aes(SDQ_total_self,group=Gruppe))+
    geom_histogram(bins=bin_count)+xlim(0,35)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-s Total'),
  ggplot(df_sample,aes(SDQ_self_int,group=Gruppe))+
    geom_histogram(bins=bin_count)+xlim(0,20)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-s Internalizing'),
  ggplot(df_sample,aes(SDQ_self_ext,group=Gruppe))+
    geom_histogram(bins=bin_count)+xlim(0,20)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-s Externalizing'),
  
nrow=2)

dev.off()

#print to markdown
knitr::include_graphics(paste0(project_path,"/output/figures/figure_questionnaire_distribution.tiff"))

```

### CBCL and SDQ correlation

- does not suggest collapsing SDQ and CBCL/YSR reports of internalizing and externalizing

```{r cbcl_sdq_correlation}

interested_variables<-c(
  "YSR_T_GES",
  "YSR_INT",
  "YSR_EXT",
  "CBCL_T_GES",
  "CBCL_T_INT",
  "CBCL_T_EXT",
  "SDQ_total_self",
  "SDQ_self_int",
  "SDQ_self_ext",
  "SDQ_total_parent",
  "SDQ_parent_int",
  "SDQ_parent_ext"
  
)

cor_matrix<-round(cor(df_sample[,interested_variables],use = 'pairwise.complete.obs'),2)
corrplot(cor_matrix,type='upper',order='hclust',tl.cex=0.6)

df_cluster<-t(df_sample[complete.cases(df_sample[,interested_variables]),interested_variables])
df_cluster<-scale(df_cluster)
dist_matrix<-dist(df_cluster, method='euclidean')
dist_mat <- as.matrix(dist_matrix)

melted_dist <- reshape2::melt(dist_mat)
ggplot(melted_dist, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "red", high = "white") +
  theme_minimal() +
  labs(title = "Euclidean Distance Matrix", x = "", y = "")

### difference SDQ CBCL/YSR
cor.test(df_sample$YSR_INT,df_sample$SDQ_self_int)
cor.test(df_sample$CBCL_INT,df_sample$SDQ_parent_int)
cor.test(df_sample$YSR_EXT,df_sample$SDQ_self_ext)
cor.test(df_sample$CBCL_EXT,df_sample$SDQ_parent_ext)

#concordance
cor.test(df_sample$YSR_INT,df_sample$CBCL_INT)
cor.test(df_sample$YSR_EXT,df_sample$CBCL_EXT)
cor.test(df_sample$SDQ_parent_int,df_sample$SDQ_self_int)
cor.test(df_sample$SDQ_parent_ext,df_sample$SDQ_self_ext)

```

### table pupil measures descriptives

```{r table pupil descriptives}

table_data<-rbind(psych::describe(df_sample$BPS_average),
                  psych::describe(df_sample$BPS_block_1),
                  psych::describe(df_sample$BPS_block_3),
                  psych::describe(df_sample$BPS_block_4),
                  psych::describe(df_sample$BPS_block_6),
                  psych::describe(df_sample$pd_low_standard_saliencehigh_utilityhigh),
                  psych::describe(df_sample$pd_low_standard_saliencehigh_utilitylow),
                  psych::describe(df_sample$pd_low_standard_saliencelow_utilityhigh),
                  psych::describe(df_sample$pd_low_standard_saliencelow_utilitylow),
                  psych::describe(df_sample$SEPR_average),
                  psych::describe(df_sample$SEPR_Oddball_before_forward),
                  psych::describe(df_sample$SEPR_Oddball_after_forward),
                  psych::describe(df_sample$rpd_oddball_saliencehigh_utilityhigh),
                  psych::describe(df_sample$rpd_oddball_saliencehigh_utilitylow),
                  psych::describe(df_sample$rpd_oddball_saliencelow_utilityhigh),
                  psych::describe(df_sample$rpd_oddball_saliencelow_utilitylow)
                  )


table_data<-table_data[,c('n','mean','sd','min','max')]
table_data<-round(table_data,2)

row.names(table_data)<-c('BPS average','BPS auditory oddball before task','BPS auditory oddball before manipulaiton','BPS auditory oddball after manipulation','BPS auditory oddball after task',
                         'BPS visual oddball ++','BPS visual oddball +-','BPS visual oddball -+','BPS visual oddball --',
                         'SEPR average','SEPR auditory oddball before manipulation','SEPR auditory oddball after manipulation',
                         'SEPR viusal oddball ++','SEPR viusal oddball +-','SEPR viusal oddball -+','SEPR viusal oddball --')


measure<-c('BPS (in mm)','','','','','','','','',
           'SEPR (in mm)','','','','','','')


task<-c('','auditory','auditory','auditory','auditory','visual','visual','visual','visual',
        '','auditory','auditory','visual','visual','visual','visual')

condition<-c('','before task','before manipulation','after manipulation','after task','salience high + utility high','salience high + utility low','salience low + utility high','salience low + utility low',
            '','before manipulation','after manipulation','salience high + utility high','salience high + utility low','salience low + utility high','salience low + utility low')


table_data<-cbind(measure,task,condition,table_data)

table_sample<-table_data %>%
  kbl(caption = "Table 1. Description of pupillary measures", row.names = F) %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  footnote(general = "++ = salience high, utility high, +- = salience high, utility low, -+ = salience low, utility high, -- = salience low, utility low")

table_sample

save_kable(table_sample, file=paste0(project_path,'/output/tables/tablex_pupildescription.html'))

```

### figure pupil measure and bio samples

```{r figure BPS SEPR}

tiff(file=paste0(project_path,"/output/figures/figure_pupillometry_biosamples.tiff"), 
     width=8, height=5.5, units="in", res=600, compression='lzw') 

composite_bins<-12

grid.arrange(

ggplot(df_sample,aes(x=BPS_average))+
  geom_histogram(bins=composite_bins)+xlab('BPS: baseline pupil size [mm]')+ylim(0,20),
ggplot(df_sample,aes(x=SEPR_average))+
  geom_histogram(bins=composite_bins)+xlab('SEPR: stimulus-evoked pupillary response [mm]')+ylim(0,20),
ggplot(df_sample,aes(x=hair_cortisol_pg_mg_log))+
  geom_histogram(bins=composite_bins)+xlab('Hair cortisol concentration [log10(pg/mg)]'),
ggplot(df_sample,aes(x=saliva_change_log))+
  geom_histogram(bins=composite_bins)+xlab('Saliva cortisol concentration change [log10(nmol/l)]'),
ncol=2
)

dev.off()

#print to markdown
knitr::include_graphics(paste0(project_path,"/output/figures/figure_pupillometry_biosamples.tiff"))

```

### covariate analysis

- female gender associated with higher self-rated psychopathology
- autism diagnosis with higher parent-rated psychopathology
- higher age with lower BPS (total or for most blocks)
- hair cortisol is higher for lower hair mass
- saliva cortisol after is lower for later time of day

- saliva cortisol before NOT influenced by time of day
- SEPR (auditory or visual oddball) is not influenced by covariates

```{r covariate analysis}

#scale numeric variables
df_scaled<-as.data.frame(lapply(df_sample, function(x) {if (is.numeric(x)) scale(x) else x}))

## effects on psychopathology
lm<-lm(YSR_GES~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on self-reported psychopathology (YSR).",
           figure_name = 'covariate_effects_on_selfreport_psychopathology_YSR')

lm<-lm(SDQ_total_self~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on self-reported psychopathology (SDQ-s).",
           figure_name = 'covariate_effects_on_selfreport_psychopathology_SDQs')

lm<-lm(CBCL_GES~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm,
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on caregiver-reported psychopathology (CBCL).",
           figure_name = 'covariate_effects_on_caregiverreport_psychopathology_CBCL')

lm<-lm(SDQ_total_parent~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm,
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on caregiver-reported psychopathology (SDQ-p).",
           figure_name = 'covariate_effects_on_caregiverreport_psychopathology_SDQp')

## effects on pupillary response
lm<-lm(BPS_average~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on baseline pupil size (BPS).",
           figure_name = 'covariate_effects_on_BPS')


  lm<-lm(BPS_block_1~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)

  lm<-lm(BPS_block_3~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)

  lm<-lm(BPS_block_5~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)

  lm<-lm(BPS_block_6~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)


lm<-lm(SEPR_average~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on stimulus-evoked pupillary response (SEPR).",
           figure_name = 'covariate_effects_on_SEPR')

  lm<-lm(SEPR_Oddball_before_forward~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)

  lm<-lm(SEPR_Oddball_after_forward~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)

  lm<-lm(rpd_oddball_saliencelow_utilitylow~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)

  lm<-lm(rpd_oddball_saliencehigh_utilityhigh~gender+age+iq+autism_boolean,data=df_scaled)
  analyze_lm(lm,log_backtransform = F)


##effects on bio samples
lm<-lm(saliva_Average.1_log~gender+age+iq+BMI+autism_boolean+saliva_CV.2+saliva_Time.1,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','BMI','autism diagnosis','measurement variation index','time of day','Residuals'),
           caption_label = "Covariate Effects on saliva cortisol concentration (SCC) before the assessment.",
           figure_name = 'covariate_effects_on_SCC_before')

lm<-lm(saliva_Average.2_log~gender+age+iq+BMI+autism_boolean+saliva_CV.2+saliva_Time.2,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','BMI','autism diagnosis','measurement variation index','time of day','Residuals'),
           caption_label = "Covariate Effects on saliva cortisol concentration (SCC) change.",
           figure_name = 'covariate_effects_on_SCC')

lm<-lm(hair_cortisol_pg_mg_log~gender+age+iq+BMI+autism_boolean+hair_mass_mg,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','BMI','autism diagnosis','hair mass','Residuals'),
           caption_label = "Covariate Effects on hair cortisol concentration (HCC).",
           figure_name = 'covariate_effects_on_HCC')


```

### correlation table cortisol and pupil measures

```{r corr table cortisol pupil measures psychopathology}

interested_variables<-c(
  "saliva_Average.1_log",
  "saliva_Average.2_log",
  "saliva_change",
  "hair_cortisol_pg_mg",
  "BPS_average", 
  #"BPS_block_1",
  #"BPS_block_2",
  "BPS_block_3",
  "BPS_block_4",
  #"BPS_block_5",
  "BPS_block_6",
  #"BPS_Standard_before_forward",
  #"BPS_Standard_before_reverse",
  #"BPS_Standard_after_forward",
  #"BPS_Standard_after_reverse",
  #"pd_low_oddball_saliencehigh_utilityhigh",
  #"pd_low_oddball_saliencehigh_utilitylow",
  #"pd_low_oddball_saliencelow_utilityhigh",
  #"pd_low_oddball_saliencelow_utilitylow",
  "SEPR_average",
  "SEPR_Oddball_before_forward",
  #"SEPR_Oddball_before_reverse",
  "SEPR_Oddball_after_forward",
  #"SEPR_Oddball_after_reverse",
  "rpd_oddball_saliencehigh_utilityhigh",
  #"rpd_oddball_saliencehigh_utilitylow",
  #"rpd_oddball_saliencelow_utilityhigh",
  "rpd_oddball_saliencelow_utilitylow",
  "YSR_T_GES",
  "YSR_T_INT",
  "YSR_T_EXT",
  "SDQ_total_self",
  "SDQ_self_int",
  "SDQ_self_ext"
  # "CBCL_T_GES",
  # "CBCL_T_INT",
  # "CBCL_T_EXT",
  # "SDQ_total_parent",
  # "SDQ_parent_int",
  # "SDQ_parent_ext"
)

data_subset<-df_sample[,interested_variables]
cor_matrix<-round(cor(data_subset,use = 'pairwise.complete.obs'),2)

# Compute p-value matrix
p_matrix <- matrix(NA, nrow = ncol(data_subset), ncol = ncol(data_subset))
rownames(p_matrix) <- colnames(p_matrix) <- colnames(data_subset)

for (i in 1:ncol(data_subset)) {
  for (j in 1:ncol(data_subset)) {
    if (i == j) {
      p_matrix[i, j] <- 1  # p = 1 for self-correlation
    } else if (i < j) {
      # Two-tailed test
      p_val <- cor.test(data_subset[[i]], data_subset[[j]], method = "pearson")$p.value
      p_matrix[i, j] <- p_val
      p_matrix[j, i] <- p_val  # symmetric
    }
  }
}

# Optionally, format p-values for readability (e.g., star notation or rounding)
p_matrix_rounded <- round(p_matrix, 3)

corrplot(
  cor_matrix, 
  type = "upper",
  method = 'color',
  tl.cex = 0.8, 
  tl.col = "black",
  p.mat = p_matrix_rounded,   # your p-value matrix
  sig.level = 0.05,           # significance threshold
  insig = "blank"              # show non-significant as p-chars or leave blank
)

```

### correlation pupillometry measures

```{r correlation pupillometry measures}

interested_variables<-c(
  "BPS_average", 
  "BPS_block_1",
  "BPS_block_2",
  "BPS_block_3",
  "BPS_block_4",
  "BPS_block_5",
  "BPS_block_6",
  "BPS_Standard_before_forward",
  "BPS_Standard_before_reverse",
  "BPS_Standard_after_forward",
  "BPS_Standard_after_reverse",
  "pd_low_oddball_saliencehigh_utilityhigh",
  "pd_low_oddball_saliencehigh_utilitylow",
  "pd_low_oddball_saliencelow_utilityhigh",
  "pd_low_oddball_saliencelow_utilitylow",
  "SEPR_average",
  "SEPR_Oddball_before_forward",
  "SEPR_Oddball_before_reverse",
  "SEPR_Oddball_after_forward",
  "SEPR_Oddball_after_reverse",
  "rpd_oddball_saliencehigh_utilityhigh",
  "rpd_oddball_saliencehigh_utilitylow",
  "rpd_oddball_saliencelow_utilityhigh",
  "rpd_oddball_saliencelow_utilitylow"
)

cor_matrix<-round(cor(df_sample[,interested_variables],use = 'pairwise.complete.obs'),2)
corrplot(cor_matrix,type='upper',method='color',tl.cex=0.8, tl.col='black')

```

### EFA key measures

```{r EFA measures}

interested_variables<-c(
  "saliva_Average.1",
  "saliva_Average.2",
  "hair_cortisol_pg_mg",
  "BPS_block_1",
  "BPS_block_2",
  "BPS_block_3",
  "BPS_block_4",
  "BPS_block_5",
  "BPS_block_6",
  "BPS_Standard_before_forward",
  "BPS_Standard_before_reverse",
  "BPS_Standard_after_forward",
  "BPS_Standard_after_reverse",
  "pd_low_oddball_saliencehigh_utilityhigh",
  "pd_low_oddball_saliencehigh_utilitylow",
  "pd_low_oddball_saliencelow_utilityhigh",
  "pd_low_oddball_saliencelow_utilitylow",
  "SEPR_Oddball_before_forward",
  "SEPR_Oddball_before_reverse",
  "SEPR_Oddball_after_forward",
  "SEPR_Oddball_after_reverse",
  "rpd_oddball_saliencehigh_utilityhigh",
  "rpd_oddball_saliencehigh_utilitylow",
  "rpd_oddball_saliencelow_utilityhigh",
  "rpd_oddball_saliencelow_utilitylow",
  "YSR_T_GES",
  "YSR_T_INT",
  "YSR_T_EXT",
  "SDQ_total_self",
  "SDQ_self_int",
  "SDQ_self_ext",
  "CBCL_T_GES",
  "CBCL_T_INT",
  "CBCL_T_EXT",
  "SDQ_total_parent",
  "SDQ_parent_int",
  "SDQ_parent_ext"
)

# Subset data and remove rows with missing values
pca_data <- df_sample[, interested_variables]
pca_data <- na.omit(pca_data)

# Scale data (recommended for PCA)
pca_data_scaled <- scale(pca_data)

# Determine optimal number of factors (e.g., using parallel analysis)
fa_parallel <- fa.parallel(pca_data_scaled, fa = 'fa', n.iter = 100, sim = F)
print(fa_parallel)

# Check if correlation matrix is positive definite
cor_mat <- cor(pca_data_scaled)
eigen_vals <- eigen(cor_mat, only.values = TRUE)$values

if (any(eigen_vals < -1e-10)) {
  message("WARNING: Correlation matrix is NOT positive definite.")
  message("Min eigenvalue: ", round(min(eigen_vals), 4))
} else {
  message("Correlation matrix is OK (all eigenvalues ≥ 0).")
}

# Optional: small regularization to fix near-zero eigenvalues
eps <- 1e-6
cor_mat_pos <- cor_mat + diag(eps, nrow(cor_mat))

# Fit EFA with, e.g., 4 factors (as suggested by your PCA scree plot)
efa_result <- fa(r = cor_mat_pos, nfactors = 6, rotate = 'oblimin', 
                 fm = 'ml', scores = 'regression')

# Summary + loading table (with threshold for readability)
print(efa_result$loadings, cut = 0.3)

# Factor correlation matrix (since rotation is oblique)
print(efa_result$Phi, digits = 2)

```

### bivariate associations 

```{r bivariate associations}

#scale numeric variables
df_scaled<-as.data.frame(lapply(df_sample, function(x) {if (is.numeric(x)) scale(x) else x}))

names(df_sample)

lm<-lm(saliva_Average.1_log~BPS_average+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','age','gender','autism diagnosis','Residuals'),
           caption_label = "Bivariate association of baseline pupil size (BPS) and saliva cortisol concentration (SCC).",
           figure_name = 'Bivariate_assoication_BPS_SCC')

lm<-lm(saliva_Average.2_log~saliva_Average.1_log+
         age+gender+autism_boolean+saliva_Time.1,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','SCC before','age','gender','autism diagnosis','time of day','Residuals'),
           caption_label = "Bivariate association of saliva cortisol concentration (SCC) before and after assessment.",
           figure_name = 'Bivariate_assoication_SCCbefore_SCCafter')


```

### cortisol and pupillary measures explaining self-reported internalizing

- lower hair cortisol associated with higher internalizing psychopathology
- higher BPS is associated with higher internalizing
- lower SEPR associated with higher internalizing
- higher level interaction of those

```{r measures on self-reported internalizing}

lm<-lm(YSR_INT~BPS_block_3*SEPR_average*hair_cortisol_pg_mg_log*saliva_Average.1_log+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC','SEPR x SCC',
                         'HCC x SCC','BPS x SEPR x HCC','BPS x SEPR x SCC',
                         'BPS x HCC x SCC','SEPR x HCC x SCC',
                         'BPS x SEPR x HCC x SCC','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC) on self-reported Internalizing (YSR Internalizing).",
           figure_name = 'combined_effects_on_YSRint')

### R² = 0.37
emtrends(lm, ~SEPR_average, var="BPS_block_3", 
         at = list(SEPR_average = c(-1,1)))

emtrends(lm, ~hair_cortisol_pg_mg_log, var="BPS_block_3", 
         at = list(hair_cortisol_pg_mg_log = c(-1,1)))


  # #residual normality
  # res_fe<-residuals(lm)
  # hist(res_fe)
  # qqnorm(res_fe)

p<-ggplot(df_scaled, aes(x=BPS_block_3,y=YSR_INT))+  
  geom_point(size=2)+
  geom_smooth(method='lm',color='blue')+
  labs(x='arousal (BPS)',y='Internalizing (YSR)')

ggplotly(p)

p<-ggplot(df_scaled, aes(x=saliva_Average.1_log,y=YSR_INT))+  
  geom_point(size=2)+
  geom_smooth(method='lm',color='red')+
  labs(x='saliva cortisol (nmol/l)',y='Internalizing (YSR)')

ggplotly(p)

p<-ggplot(df_scaled, aes(x=hair_cortisol_pg_mg_log,y=YSR_INT))+  
  geom_point(size=2)+
  geom_smooth(method='lm',color='darkgreen')+
  labs(x='hair cortisol (pg/mL)',y='Internalizing (YSR)')

ggplotly(p)


## SDQ Internalizing
### R² = 0.33
lm<-lm(SDQ_self_int~BPS_block_3*SEPR_average*hair_cortisol_pg_mg_log*saliva_Average.1_log+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC change',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC change','SEPR x SCC change',
                         'HCC x SCC change','BPS x SEPR x HCC','BPS x SEPR x SCC change',
                         'BPS x HCC x SCC change','SEPR x HCC x SCC change',
                         'BPS x SEPR x HCC x SCC change','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Internalizing (SDQ-s Internalizing).",
           figure_name = 'combined_effects_on_SDQint')


  
```


### cortisol and pupillary measures explaining self-reported externalizing

- higher SCC is assoicated with higher Externalizing in YSR and SDQs

```{r measures on self-reported externalizing}

#YSR Ext
lm<-lm(YSR_EXT~BPS_block_3*SEPR_average*hair_cortisol_pg_mg_log*saliva_Average.1_log+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
summary(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC','SEPR x SCC',
                         'HCC x SCC','BPS x SEPR x HCC','BPS x SEPR x SCC',
                         'BPS x HCC x SCC','SEPR x HCC x SCC',
                         'BPS x SEPR x HCC x SCC','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Externalizing (YSR).",
           figure_name = 'combined_effects_on_YSRext')

### R² = 0.17
emtrends(lm, ~hair_cortisol_pg_mg_log, var="BPS_block_3", at = list(hair_cortisol_pg_mg_log = c(-1,1)))

p<-ggplot(df_scaled, aes(x=saliva_Average.1_log,y=YSR_EXT))+  
  geom_point(size=2)+
  geom_smooth(method='lm',color='orange')+
  labs(x='saliva cortisol (nmol/L)',y='Externalizing (YSR)')

ggplotly(p)


#SDQ Ext
lm<-lm(SDQ_self_ext~BPS_block_3*SEPR_average*hair_cortisol_pg_mg_log*saliva_Average.1_log+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
### R² = 0.06

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC change',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC change','SEPR x SCC change',
                         'HCC x SCC change','BPS x SEPR x HCC','BPS x SEPR x SCC change',
                         'BPS x HCC x SCC change','SEPR x HCC x SCC change',
                         'BPS x SEPR x HCC x SCC change','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Externalizing (SDQ-s).",
           figure_name = 'combined_effects_on_SDQext')

emtrends(lm, ~BPS_block_3, var="SEPR_average", at = list(BPS_block_3 = c(-2,-1,0,1,2)))


# df_scaled$hair_split<-ifelse(df_scaled$hair_cortisol_pg_mg>=0,'above average hair HCC','below average HCC')
# df_scaled$saliva_split<-ifelse(df_scaled$saliva_Average.1_log>=0,'above average SCC','below average SCC')
# ggplot(df_scaled[!is.na(df_scaled$hair_split) & !is.na(df_scaled$saliva_split),],
#        aes(x=SEPR_average,y=YSR_EXT))+
#   geom_point(size=2)+
#   geom_smooth(method='lm',color='darkgrey')+
#   facet_wrap(~hair_split+saliva_split)+
#   theme(legend.position = 'none')


```

### classification

- above change classification based on pupillometry and cortisol measures

```{r classification, warnings=F}
set.seed(333)

split_ratio<-0.2

df_sample$YSR_INT_CUTOFF<-ifelse(df_sample$YSR_T_INT>=65,T,F)
df_sample$YSR_EXT_CUTOFF<-ifelse(df_sample$YSR_T_EXT>=65,T,F)

# Ensure the target variable is a factor
df_sample$YSR_INT_CUTOFF <- as.factor(df_sample$YSR_INT_CUTOFF)
df_sample$YSR_EXT_CUTOFF <- as.factor(df_sample$YSR_EXT_CUTOFF)


table(df_sample$YSR_INT_CUTOFF)
table(df_sample$YSR_EXT_CUTOFF)

table(df_sample$YSR_INT_CUTOFF,df_sample$YSR_EXT_CUTOFF)


#complete data
df_sample_complete<-df_sample[complete.cases(df_sample$BPS_block_3) & 
                  complete.cases(df_sample$SEPR_average) & 
                  complete.cases(df_sample$hair_cortisol_pg_mg_log) & 
                  complete.cases(df_sample$saliva_Average.1_log) & complete.cases(df_sample$YSR_INT_CUTOFF),]


#bootstrapping function - cortisol
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_INT_CUTOFF ~ hair_cortisol_pg_mg_log*saliva_Average.1_log, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}

# bootstrapping
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("cortisol AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


#bootstrapping function - pupillometry
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_INT_CUTOFF ~ BPS_block_3*SEPR_average, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}

# Combine training data and testData for bootstrapping
# (You can also bootstrap on testData if preferred)
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("pupillometry AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


#bootstrapping function - pupillometry + hair cortisol
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_INT_CUTOFF ~ BPS_block_3*SEPR_average*hair_cortisol_pg_mg_log*saliva_Average.1_log, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}

# Combine training data and testData for bootstrapping
# (You can also bootstrap on testData if preferred)
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("pupillometry + hair cortisol AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


####FOR EXTERNALIZING

# Bootstrapping function to compute AUC
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_EXT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_EXT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_EXT_CUTOFF ~ BPS_block_3*SEPR_average*hair_cortisol_pg_mg_log*saliva_Average.1_log, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_EXT_CUTOFF, probs)
  return(auc(roc_obj))
}

# Combine training data and testData for bootstrapping
# (You can also bootstrap on testData if preferred)
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("pupillometry + cortisol AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


```

