---
title: "data_analysis_biosamples_psychopathology"
author: "Nico Bast"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    theme: cerulean
    code_folding: hide
  word_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#required packages
require(openxlsx) #read xls
require(ggplot2)
require(gridExtra)
require(effectsize) #standardized coefficients of LM
require(kableExtra)
require(corrplot)
require(car) # ANOVA type=3
require(emmeans) # post hoc
require(caret) #classification
require(pROC) #classification
require(boot) #classification


#set ggplot theme
theme_set(theme_bw())

#custom functions

#customaized ANOVA output
analyze_lm<-function(x,log_backtransform=F){
  
  model_summary<-summary(x)
  print(paste('adj. R²:',round(model_summary$adj.r.squared,2)))
  
  print(car::Anova(x,type=3))
  return(cbind(round(summary(x)$coefficients[,1],2),round(confint(x),2)))
  
  if(log_backtransform){
  return(cbind(round(10^summary(x)$coefficients[,1],2),round(10^confint(x),2)))
  }
  
}

#create tables for supplements
save_model<-function(lm_output=lm,
                     var_names=c(attr(lm$coefficients,'names'),'Residuals'),
                     caption_label,
                     figure_name){

  
#table 1 
table_data<-round(car::Anova(lm_output,type=3),3)
table_data<-cbind(var_names,table_data)

table_output<-table_data %>%
  kbl(caption = paste0('Model summary - ',caption_label),
      col.names = c('','Sum Sq','DF','F-value','p-value'),
      row.names = F) %>%
  kable_classic(full_width = F, html_font = "Cambria")

save_kable(table_output, file=paste0(project_path,'/output/supplements/',figure_name,'_1.html'))

#table 2
table_data2<-data.frame(cbind(round(summary(lm_output)$coefficients[,1],2),round(confint(lm_output),2)))
table_data2<-cbind(head(var_names,-1),table_data2)

table_output<-table_data2 %>%
  kbl(caption = paste0('Effect estimates - ',caption_label),
      col.names = c('','estimate (beta)','lower bound (2.5%)','upper bound (97.5%)'),
      row.names = F) %>%
  kable_classic(full_width = F, html_font = "Cambria")

save_kable(table_output, file=paste0(project_path,'/output/supplements/',figure_name,'_2.html'))

}

```

# SEGA biosamples psychopathology - stress

Aim: relate biosample data (hair and saliva) and pupillometry (BPS and SEPR) to transdiagnostic psychopathology

## data assembly

- combine different data sources and calculate core variables

### read data

- utilizes a master file of the SEGA data

```{r read data}

#local data path - masterfile
root_path<-'C:/Users/nico/Nextcloud/project_sega'

#remote data path - WINET
data_path<-'//192.168.88.212/daten/KJP_Studien/Studie_SEGA'

#project folder
project_path<-"C:/Users/nico/Nextcloud/Paper_SEGA_stress"


#initial master file - read after row 2
df<-read.xlsx(paste0(root_path,'/data/master_file_SEGA_data.xlsx'),na.strings = c("NA","-"),detectDates = T)

#experimental data from later master file
df_exp<-read.xlsx(paste0(root_path,'/data/master_file_SEGA_data_Sept2025.xlsx'),startRow = 3)
df_exp<-df_exp[,c(1,3,57:88)] 

#eye-tracking - trial-based visual oddball
df_trial_visualoddball<-readRDS(paste0(root_path,'/data/preprocessed_df_trial.rds'))

#full data but without SDQ data
df_handdyn<-read.xlsx(paste0(root_path,'/data/master_file_SEGA_Oct2025.xlsx'),na.strings = c("NA","-"),detectDates = T)

#comorbidites
df_comorbidity<-read.xlsx(paste0(root_path,'/data/comorbidities_SEGA.xlsx'),detectDates = T)

#read CBCL
df_CBCL<-read.xlsx(paste0(data_path,"/5_Versuchsdaten/Datenbank Export/2026_08_13/CBCL_4_18.xlsx"))


```

```{r cbcl factor scores}

cbcl_items<-df_CBCL[,c(6:125)]
cbcl_items[cbcl_items == 2] <- 1

attach(cbcl_items)
cbcl_P_f<-
  0.535 * a14 +
  0.339 * a29 +
  0.383 * a30 +
  0.510 * a31 +
  0.175 * a32 +
  0.665 * a33 +
  0.571 * a35 +
  0.607 * a45 +
  0.501 * a50 +
  0.343 * a52 +
  0.398 * a71 +
  0.591 * a91 +
  0.488 * a112 +
  0.372 * a42 +
  0.595 * a65 +
  0.579 * a69 +
  0.197 * a75 +
  0.506 * a102 +
  0.652 * a103 +
  0.384 * a111 +
  0.423 * a47 +
  0.283 * a49 +
  0.308 * a51 +
  0.397 * a54 +
  0.323 * a56a +
  0.317 * a56b +
  0.373 * a56c +
  0.228 * a56d +
  0.234 * a56e +
  0.183 * a56f +
  0.161 * a56g +
  0.669 * a26 +
  0.565 * a39 +
  0.655 * a43 +
  0.440 * a63 +
  0.502 * a67 +
  0.402 * a73 +
  0.565 * a81 +
  0.576 * a82 +
  0.604 * a90 +
  0.526 * a96 +
  0.389 * a101 +
  0.410 * a105 +
  0.463 * a106 +
  0.647 * a3 +
  0.644 * a16 +
  0.673 * a19 +
  0.646 * a20 +
  0.666 * a21 +
  0.715 * a22 +
  0.601 * a23 +
  0.697 * a37 +
  0.617 * a57 +
  0.669 * a68 +
  0.777 * a86 +
  0.745 * a87 +
  0.727 * a88 +
  0.638 * a89 +
  0.601 * a94 +
  0.720 * a95 +
  0.564 * a97 +
  0.644 * a104 +
  0.495 * a1 +
  0.577 * a8 +
  0.518 * a10 +
  0.519 * a13 +
  0.524 * a17 +
  0.670 * a41 +
  0.578 * a61 +
  0.499 * a80

cbcl_INT_f <-
  0.130 * a14 +
  0.310 * a29 +
  0.286 * a30 +
  0.322 * a31 +
  0.198 * a32 +
  0.177 * a33 +
  0.314 * a35 +
  0.332 * a45 +
  0.410 * a50 +
  0.506 * a52 +
  0.381 * a71 +
  (-0.020) * a91 +
  0.431 * a112 +
  0.226 * a42 +
  0.217 * a65 +
  0.145 * a69 +
  0.392 * a75 +
  0.336 * a102 +
  0.308 * a103 +
  0.407 * a111 +
  0.276 * a47 +
  0.303 * a49 +
  0.481 * a51 + 
  0.384 * a54 +
  0.432 * a56a +
  0.491 * a56b +
  0.633 * a56c +
  0.091 * a56d +
  0.262 * a56e +
  0.676 * a56f +
  0.494 * a56g

cbcl_EXT_f <-
  0.157 * a26 +
  0.388 * a39 +
  0.260 * a43 +
  0.163 * a63 +
  0.408 * a67 +
  0.315 * a72 +
  0.437 * a81 +
  0.533 * a82 +
  0.204 * a90 +
  0.159 * a96 +
  0.546 *a101 + 
  0.521 * a105 +
  0.486 * a106 +
  (-0.069) * a3 +
  0.191 * a16 +
  (-0.005) * a19 +
  0.363 * a20 +
  0.390 * a21 +
  0.134 * a22 +
  0.388 * a23 +
  0.305 * a37 +
  0.291 * a57 +
  0.055 * a68 +
  (-0.225) * a86 +
  (-0.254) * a87 +
  (-0.075) * a88 +
  0.188 * a89 +
  0.148 * a94 +
  0.115 * a95 +
  0.313 * a97 +
  0.073 * a104

cbcl_ATT_f<-
0.300*a1 +
0.644*a8 +
0.493*a10 +
0.331*a13 +
0.399*a17 +
0.244*a41 +
0.211*a61 +
0.256*a80


hist(cbcl_P_f)
hist(df_CBCL$CBCL_GES)
cor.test(cbcl_P_f,df_CBCL$CBCL_GES,method='pearson',exact=F)

hist(cbcl_INT_f)
hist(df_CBCL$CBCL_INT)
cor.test(cbcl_INT_f,df_CBCL$CBCL_INT,method='pearson',exact=F)

hist(cbcl_EXT_f)
hist(df_CBCL$CBCL_EXT)
cor.test(cbcl_EXT_f,df_CBCL$CBCL_EXT,method='pearson',exact=F)

hist(cbcl_ATT_f)
cor.test(cbcl_ATT_f,df_CBCL$CBCL_AP,method='pearson',exact=F)


df_CBCL_f<-data.frame(ID_Bado=df_CBCL$ID_Bado,cbcl_P_f,cbcl_INT_f,cbcl_EXT_f,cbcl_ATT_f)

detach(cbcl_items)

```



### rename variables

```{r rename variables}

names(df)

#saliva samples renaming
names(df)[31:40]<-paste0('saliva_',names(df)[31:40])

names(df_handdyn)[42:45]<-c('saliva_Value.1','saliva_Value.2','saliva_CV.1','saliva_CV.2')

#hair renaming
names(df)[26]<-'hair_mass_mg'
names(df)[27]<-'hair_comments'
names(df)[28]<-'hair_cortisol_pg_mg'
names(df)[29]<-'hair_cortisone_pg_mg'

names(df_handdyn)[38:40]<-c('hair_mass_mg','hair_cortisol_pg_mg','hair_cortisone_pg_mg')

#SDQ
names(df)[41:46]<-c('SDQ_emo','SDQ_beh','SDQ_hyp','SDQ_beh_peers','SDQ_pros','SDQ_total')
names(df)[47:52]<-c('SDQ_emo','SDQ_beh','SDQ_hyp','SDQ_beh_peers','SDQ_pros','SDQ_total')

names(df)[41:46]<-paste0(names(df)[41:46],'_parent')
names(df)[47:52]<-paste0(names(df)[47:52],'_self')

#remove special characters
names(df)<-gsub('\\[','',names(df))
names(df)<-gsub('\\]','',names(df))
names(df)<-gsub('\\%','',names(df))

```

### merge data

```{r merge}

#merge demographics and experimental data
df<-merge(df,df_exp,by.x='SEGA_ID',by.y='ID_Studie')

#merge CBCL factor scores - 4 participants unmatched
df<-merge(df,df_CBCL_f,by.x='ID-BADO',by.y='ID_Bado')


```

### calculate age

```{r add age}

hist(df_handdyn$date_data_collection,20)
hist(df_handdyn$birthday,20)

#merge age
df_handdyn$age<-as.numeric(with(df_handdyn,date_data_collection-birthday)/365)
df$merger_id<-as.numeric(substr(df$SEGA_ID,6,8))
df_age<-df_handdyn[,names(df_handdyn) %in% c('sega_id','age')]

df<-merge(df,df_age,by.x='merger_id',by.y='sega_id')

```

### add comorbidity data

```{r add comorbidity}

df<-merge(df,df_comorbidity,by.x='merger_id',by.y='SEGA_ID')

```

## data manipulation

### create saliva change variable

```{r}

df$saliva_change<-df$saliva_Average.2-df$saliva_Average.1

```

### recode group variable

```{r}

table(df$group)

df$group<-ifelse(df$Gruppe %in% c('AUT','ASD'),'ASD',df$Gruppe)
df$autism_boolean<-ifelse(df$group=='ASD',T,F)

```

### logarithmize bio samples

```{r}

df$saliva_Average.1_log<-log10(df$saliva_Average.1)
df$saliva_Average.2_log<-log10(df$saliva_Average.2)
df$hair_cortisol_pg_mg_log<-log10(df$hair_cortisol_pg_mg)

df$saliva_change_log<-df$saliva_Average.2_log-df$saliva_Average.1_log
df$saliva_change<-df$saliva_Average.2-df$saliva_Average.1

```

### IQ estimate

```{r}

df$iq<-rowMeans(df[,names(df) %in% c('iq_mt','iq_gf','iq_mz','iq_wt')],na.rm=T)
hist(df$iq)

```

### estimate BMI

```{r}

df$BMI<-with(df,Gewicht.kg/(Groeße.cm/100*Groeße.cm/100))

```

### pupil measure averages

```{r}

df$BPS_average<-(df$pd_low_oddball_saliencehigh_utilityhigh+
  df$pd_low_oddball_saliencehigh_utilitylow+
  df$pd_low_oddball_saliencelow_utilityhigh+
  df$pd_low_oddball_saliencelow_utilitylow+
  df$BPS_Oddball_before_forward+
  df$BPS_Oddball_after_forward+
  df$BPS_Oddball_before_reverse+
  df$BPS_Oddball_after_reverse)/8


df$SEPR_average<-(df$rpd_oddball_saliencehigh_utilityhigh+
                    df$rpd_oddball_saliencehigh_utilitylow+
                    df$rpd_oddball_saliencelow_utilityhigh+
                    df$rpd_oddball_saliencelow_utilitylow+
                    df$SEPR_Oddball_after_forward+
                    df$SEPR_Oddball_after_reverse+
                    df$SEPR_Oddball_before_forward+
                    df$SEPR_Oddball_before_reverse)/8

hist(df$BPS_average,20)
hist(df$SEPR_average,20)  

```

### pupil size changes

```{r}

df$SEPR_Oddball_change<-df$SEPR_Oddball_after_forward-df$SEPR_Oddball_before_forward
df$BPS_change<-df$BPS_Standard_after_forward-df$BPS_Standard_before_forward

```

### SDQ externalizing internalizing - self report

- use Germans norms
- calculate internalizing and externalizing score

```{r}

#norms
norms_emo_self <- data.frame(
  raw = 0:10,
  percentile = c(16, 38, 58, 74, 86, 92, 96, 98, 99, 100, 100)
)

norms_beh_peers_self <- data.frame(
  raw = 0:10,
  percentile = c(16, 43, 68, 84, 93, 97, 99, 100, 100, 100, 100)
)

norms_beh_self <- data.frame(
  raw = 0:10,
  percentile = c(12, 43, 72, 88, 95, 98, 99, 100, 100, 100, 100)
)

norms_hyp_self <- data.frame(
  raw = 0:10,
  percentile = c(6, 16, 31, 49, 66, 83, 91, 97, 99, 100, 100)
)

#translate raw to percentiles based on norms
df$pcts_emo_self  <- norms_emo_self$percentile[match(df$SDQ_emo_self, norms_emo_self$raw) ]
df$pcts_beh_peers_self  <- norms_beh_peers_self$percentile[match(df$SDQ_beh_peers_self, norms_beh_peers_self$raw) ]
df$pcts_beh_self  <- norms_beh_self$percentile[match(df$SDQ_beh_self, norms_beh_self$raw) ]
df$pcts_hyp_self  <- norms_hyp_self$percentile[match(df$SDQ_hyp_self, norms_hyp_self$raw) ]

#PERCENTILES: internalizing
df$SDQ_self_int_pcts<-(df$pcts_emo_self+df$pcts_beh_peers_self)/2
df$SDQ_self_ext_pcts<-(df$pcts_beh_self+df$pcts_hyp_self)/2

#RAW: internalizing 
df$SDQ_self_int<-df$SDQ_emo_self+df$SDQ_beh_peers_self
df$SDQ_self_ext<-df$SDQ_beh_self+df$SDQ_hyp_self

grid.arrange(
  ggplot(df,aes(SDQ_self_int_pcts))+geom_histogram(),
  ggplot(df,aes(SDQ_self_ext_pcts))+geom_histogram(),
  ggplot(df,aes(SDQ_self_int))+geom_histogram(),
  ggplot(df,aes(SDQ_self_ext))+geom_histogram()
)

```

### SDQ externalizing internalizing - parent report

- use Germans norms
- calculate internalizing and externalizing score

```{r}

#norms
norms_emo_parent <- data.frame(
  raw = 0:10,
  percentile = c(16, 38, 58, 74, 86, 92, 96, 98, 99, 100, 100)
)

norms_beh_peers_parent <- data.frame(
  raw = 0:10,
  percentile = c(16, 43, 68, 84, 93, 97, 99, 100, 100, 100, 100)
)

norms_beh_parent <- data.frame(
  raw = 0:10,
  percentile = c(12, 43, 72, 88, 95, 98, 99, 100, 100, 100, 100)
)

norms_hyp_parent <- data.frame(
  raw = 0:10,
  percentile = c(6, 16, 31, 49, 66, 83, 91, 97, 99, 100, 100)
)

#translate raw to percentiles based on norms
df$pcts_emo_parent  <- norms_emo_parent$percentile[match(df$SDQ_emo_parent, norms_emo_parent$raw) ]
df$pcts_beh_peers_parent  <- norms_beh_peers_parent$percentile[match(df$SDQ_beh_peers_parent, norms_beh_peers_parent$raw) ]
df$pcts_beh_parent  <- norms_beh_parent$percentile[match(df$SDQ_beh_parent, norms_beh_parent$raw) ]
df$pcts_hyp_parent  <- norms_hyp_parent$percentile[match(df$SDQ_hyp_parent, norms_hyp_parent$raw) ]

#PERCENTILES: internalizing
df$SDQ_parent_int_pcts<-df$pcts_emo_parent+df$pcts_beh_peers_parent
df$SDQ_parent_ext_pcts<-df$pcts_beh_parent+df$pcts_hyp_parent

#RAW: internalizing 
df$SDQ_parent_int<-df$SDQ_emo_parent+df$SDQ_beh_peers_parent
df$SDQ_parent_ext<-df$SDQ_beh_parent+df$SDQ_hyp_parent

```

### global externalizing internalizing

```{r}

cor.test(df$SDQ_self_int_pcts,df$YSR_T_INT)
cor.test(df$SDQ_self_ext_pcts,df$YSR_T_EXT)


df$INT_self<-rowMeans(scale(df$SDQ_self_int)+scale(df$YSR_T_INT))
df$EXT_self<-rowMeans(scale(df$SDQ_self_ext)+scale(df$YSR_T_EXT))

df$INT_parent<-rowMeans(scale(df$SDQ_parent_int)+scale(df$CBCL_T_INT))
df$EXT_parent<-rowMeans(scale(df$SDQ_parent_ext)+scale(df$CBCL_T_EXT))

hist(df$INT_self,20)
hist(df$EXT_self,20)
hist(df$INT_parent,20)
hist(df$EXT_parent,20)

```

### outlier exclusion 

```{r}

#questionnaires
hist(df$YSR_GES)
hist(df$YSR_T_GES)

#correct typo
df$YSR_GES[df$YSR_T_GES>100 & !is.na(df$YSR_T_GES)]
df$YSR_GES[df$YSR_T_GES<25 & !is.na(df$YSR_T_GES)]

df$YSR_T_GES[which(df$YSR_T_GES>100)]<-48
df$YSR_T_GES[which(df$YSR_T_GES<25)]<-89

##--> other scales are okay
```

## data quality

### missing data

-complete data for CBCL
-more missings in SDQ

```{r missing data}

CBCL<-with(df,table(!is.na(CBCL_T_GES)))
YSR<-with(df,table(!is.na(YSR_T_GES)))

SDQ_self<-with(df,table(!is.na(SDQ_total_self)))
SDQ_parent<-with(df,table(!is.na(SDQ_total_parent)))

hair_cortisol<-with(df,table(!is.na(hair_cortisol_pg_mg)))
saliva_cortisol_measure1<-with(df,table(!is.na(saliva_Average.1)))
saliva_cortisol_measure2<-with(df,table(!is.na(saliva_Average.2)))

BPS_average_mis<-with(df,table(!is.na(BPS_average)))
SEPR_average_mis<-with(df,table(!is.na(SEPR_average)))

missing_table<-rbind(CBCL,YSR,SDQ_self,SDQ_parent,hair_cortisol,saliva_cortisol_measure1,saliva_cortisol_measure2,
                     BPS_average_mis,SEPR_average_mis)

missing_table %>%
  kbl(caption = "Available data for data modalities") %>%
  kable_classic(full_width = F, html_font = "Cambria")

with(df,table(!is.na(YSR_T_GES),!is.na(hair_cortisol_pg_mg)))
with(df,table(!is.na(SDQ_total_parent),!is.na(hair_cortisol_pg_mg)))
with(df,table(!is.na(hair_cortisol_pg_mg),!is.na(BPS_average)))
with(df,table(!is.na(hair_cortisol_pg_mg),!is.na(saliva_Average.1)))

```

### participants by diagnosis and available biosamples

```{r}

table(df$group[!is.na(df$hair_cortisol_pg_mg) & !is.na(df$saliva_Average.1)])

```

### bio samples data quality

#### hair cortisol

```{r hair}

ggplot(df,aes(x=hair_cortisone_pg_mg,y=hair_cortisol_pg_mg,color=hair_mass_mg))+
  geom_point()+
  labs(x='hair cortisone (pg/mg, log-scaled)', y='hair cortisol (pg/mg, log-scaled)', color='sample mass (mg)')+
  geom_smooth(color='grey',method='lm')+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

```

#### saliva cortisol

```{r saliva}

hist(df$saliva_Average.1)
hist(df$saliva_Average.2)

ggplot(df,aes(y=saliva_Average.1,x=saliva_CV.1))+
  geom_point()+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

ggplot(df,aes(y=saliva_Average.2,x=saliva_CV.2))+
  geom_point()+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

ggplot(df,aes(x=saliva_Average.2-saliva_Average.1))+
  geom_histogram()+
  geom_vline(xintercept=0,linetype=2)+
  labs(title='saliva cortisol reduction within experiment (pg/mg)')

ggplot(df,aes(x=saliva_Average.1,y=saliva_Average.2,color=saliva_CV.1))+
  geom_point()+
  geom_smooth(color='grey',method='lm')+
  #labs(x='hair cortisone (pg/mg, log-scaled)', y='hair cortisol (pg/mg, log-scaled)', color='sample mass (mg)')+
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

```

##### intra- inter-assay coefficient of variation (CV)

```{r intra-assay CV}

psych::describe(df$saliva_CV.1)
psych::describe(df$saliva_CV.2)

```

### bio samples

- suggests to log-transform variables in analysis

```{r data inspection bio samples}

ggplot(df,aes(hair_cortisol_pg_mg))+geom_histogram()
ggplot(df,aes(saliva_Average.1))+geom_histogram()
ggplot(df,aes(saliva_Average.2))+geom_histogram()
ggplot(df,aes(saliva_change))+geom_histogram()

ggplot(df,aes(x=hair_cortisol_pg_mg,y=saliva_change))+geom_point()+
  scale_x_continuous(trans='log10')+
  scale_y_continuous(trans='log10')+
  geom_smooth(method='lm')

```

### covariates biosamples - influence of bio sample characteristics

- considering: time of day, saliva measurement error, BMI, age, gender
- higher saliva measurement error associated with lower saliva cortisol 1
- higher age associated with higher saliva cortisol 1
- higher hair mass associated with lower hair cortisol measure

--> time of day, gender, BMI is not associated with biosamples

```{r covariates biosamples}

lm<-lm(saliva_Average.1_log~saliva_Time.1+saliva_CV.1+BMI+age+gender,df)
anova(lm)
standardize_parameters(lm)

anova(lm(saliva_Average.2_log~saliva_Time.2+saliva_CV.2+BMI+age+gender,df))

lm<-lm(hair_cortisol_pg_mg~hair_mass_mg+BMI+age+gender,df)
anova(lm)
standardize_parameters(lm)


```

### covariates questionnaires

```{r covariates questionnaires}

#YSR - self report
lm<-lm(YSR_T_GES~age+gender+iq+autism_boolean,df)
anova(lm)

lm<-lm(YSR_T_INT~age+gender+iq+autism_boolean,df)
anova(lm)

lm<-lm(YSR_T_EXT~age+gender+iq+autism_boolean,df)
anova(lm)

#CBCL - parent report
lm<-lm(CBCL_T_GES~age+gender+iq+autism_boolean,df)
anova(lm)

lm<-lm(CBCL_T_INT~age+gender+iq+autism_boolean,df)
anova(lm)

lm<-lm(CBCL_T_EXT~age+gender+iq+autism_boolean,df)
anova(lm)

#SDQ - self report
lm<-lm(SDQ_self_int~age+gender+iq+autism_boolean,df)
anova(lm)
summary(lm)

lm<-lm(SDQ_self_ext~age+gender+iq+autism_boolean,df)
anova(lm)

#SDQ - parent report
lm<-lm(SDQ_parent_int~age+gender+iq+autism_boolean,df)
anova(lm)
summary(lm)

lm<-lm(SDQ_parent_ext~age+gender+iq+autism_boolean,df)
anova(lm)
summary(lm)


```

## data analysis

### sample definition

- define sample as participants with self-reported psychopathology and hair_cortisol

```{r sample definition}

df_sample<-df[!is.na(df$YSR_T_GES) & !is.na(df$hair_cortisol_pg_mg),]

#toial sample
nrow(df_sample)

```

### comorbidities

```{r comorbidites}

#with primary diagnosis
table(df_sample$group=='CON')

#primnary diagnosis
table(df_sample$primary.diagnosis)

#comorbdiity
table(!is.na(df_sample$other.diagnoses))
table(c(df_sample$other.diagnoses,df_sample$other.diagnoses.1))

```

### sample description

```{r sample description}

table(df_sample$gender)

table_data<-rbind(psych::describe(df_sample$age),
                  psych::describe(df_sample$iq),
                  psych::describe(df_sample$YSR_T_GES),
                  psych::describe(df_sample$YSR_T_INT),
                  psych::describe(df_sample$YSR_T_EXT),
                  psych::describe(df_sample$CBCL_T_GES),
                  psych::describe(df_sample$CBCL_T_INT),
                  psych::describe(df_sample$CBCL_T_EXT),
                  psych::describe(df_sample$SDQ_total_self),
                  psych::describe(df_sample$SDQ_self_int),
                  psych::describe(df_sample$SDQ_self_ext),
                  psych::describe(df_sample$SDQ_total_parent),
                  psych::describe(df_sample$SDQ_parent_int),
                  psych::describe(df_sample$SDQ_parent_ext),
                  psych::describe(df_sample$hair_cortisol_pg_mg),
                  psych::describe(df_sample$saliva_Average.1),
                  psych::describe(df_sample$saliva_Average.2))


table_data<-table_data[,c('n','mean','sd','min','max')]
table_data<-round(table_data,2)

row.names(table_data)<-c('age','iq','YSR TOTAL (t)','YSR INT (t)','YSR EXT (t)',
                         'CBCL TOTAL (t)','CBCL INT (t)','CBCL EXT (t)',
                         'SDQ TOTAL self (sum)','SDQ INT self (sum)','SDQ EXT self (sum)',
                         'SDQ TOTAL parent (sum)','SDQ INT parent (sum)','SDQ EXT parent (sum)',
                         'hair cortisol (pg/mg)','saliva cortisol before (nmol/l)','saliva cortisol after (nmol/l)')


table_sample<-table_data %>%
  kbl(caption = "Table 1. Sample description") %>%
  kable_classic(full_width = F, html_font = "Cambria")
  

table_sample

save_kable(table_sample, file=paste0(project_path,'/output/tables/table1_sampledescription.html'))

```

### figure pupil measure and bio samples

```{r figure BPS SEPR}

tiff(file=paste0(project_path,"/output/figures/figure_pupillometry_biosamples.tiff"), 
     width=8, height=5.5, units="in", res=600, compression='lzw') 

composite_bins<-12

grid.arrange(

ggplot(df_sample,aes(x=BPS_average))+
  geom_histogram(bins=composite_bins)+xlab('BPS: baseline pupil size [mm]')+ylim(0,20),
ggplot(df_sample,aes(x=SEPR_average))+
  geom_histogram(bins=composite_bins)+xlab('SEPR: stimulus-evoked pupillary response [mm]')+ylim(0,20),
ggplot(df_sample,aes(x=hair_cortisol_pg_mg_log))+
  geom_histogram(bins=composite_bins)+xlab('Hair cortisol concentration [log10(pg/mg)]'),
ggplot(df_sample,aes(x=saliva_change_log))+
  geom_histogram(bins=composite_bins)+xlab('Saliva cortisol concentration change [log10(nmol/l)]'),
ncol=2
)

dev.off()

#print to markdown
knitr::include_graphics(paste0(project_path,"/output/figures/figure_pupillometry_biosamples.tiff"))

```


### figure questionnaire data

```{r data inspection questionnaires}

bin_count<-15

tiff(file=paste0(project_path,"/output/figures/figure_questionnaire_distribution.tiff"), 
     width=8, height=4, units="in", res=600, compression='lzw') 

#core questionnaire measures
grid.arrange(
  # ggplot(df_sample,aes(CBCL_GES,group=Gruppe))+
  #   geom_histogram(bins=bin_count)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('CBCL Total'),
  # ggplot(df_sample,aes(CBCL_INT,group=Gruppe))+
  #   geom_histogram(bins=bin_count)+ylim(0,20)+xlim(0,50)+theme(axis.title.y = element_blank())+xlab('CBCL Internalizing'),
  # ggplot(df_sample,aes(CBCL_EXT,group=Gruppe))+
  #   geom_histogram(bins=bin_count)+ylim(0,32)+xlim(0,50)+theme(axis.title.y = element_blank())+xlab('CBCL Externalizing'),
  # 
  
  ggplot(df_sample,aes(YSR_GES,group=Gruppe))+
    geom_histogram(bins=bin_count)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('YSR Total'),
  ggplot(df_sample,aes(YSR_INT,group=Gruppe))+
    geom_histogram(bins=bin_count)+ylim(0,20)+xlim(0,50)+theme(axis.title.y = element_blank())+xlab('YSR Internalizing'),
  ggplot(df_sample,aes(YSR_EXT,group=Gruppe))+
    geom_histogram(bins=bin_count)+ylim(0,32)+xlim(0,50)+theme(axis.title.y = element_blank())+xlab('YSR Externalizing'),
  
  ggplot(df_sample,aes(SDQ_total_self,group=Gruppe))+
    geom_histogram(bins=bin_count)+xlim(0,35)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-s Total'),
  ggplot(df_sample,aes(SDQ_self_int,group=Gruppe))+
    geom_histogram(bins=bin_count)+xlim(0,20)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-s Internalizing'),
  ggplot(df_sample,aes(SDQ_self_ext,group=Gruppe))+
    geom_histogram(bins=bin_count)+xlim(0,20)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-s Externalizing'),
  
  # ggplot(df_sample,aes(SDQ_total_parent,group=Gruppe))+
  #   geom_histogram(bins=bin_count)+xlim(0,35)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-p Total'),
  # ggplot(df_sample,aes(SDQ_parent_int,group=Gruppe))+
  #   geom_histogram(bins=bin_count)+xlim(0,20)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-p Internalizing'),
  # ggplot(df_sample,aes(SDQ_parent_ext,group=Gruppe))+
  #   geom_histogram(bins=bin_count)+xlim(0,20)+ylim(0,20)+theme(axis.title.y = element_blank())+xlab('SDQ-p Externalizing')
  
nrow=2)

dev.off()

#print to markdown
knitr::include_graphics(paste0(project_path,"/output/figures/figure_questionnaire_distribution.tiff"))

```



### CBCL and SDQ correlation

- does not suggest collapsing SDQ and CBCL/YSR reports of internalizing and externalizing

```{r}

interested_variables<-c(
  "YSR_T_GES",
  "YSR_INT",
  "YSR_EXT",
  "CBCL_T_GES",
  "CBCL_T_INT",
  "CBCL_T_EXT",
  "SDQ_total_self",
  "SDQ_self_int",
  "SDQ_self_ext",
  "SDQ_total_parent",
  "SDQ_parent_int",
  "SDQ_parent_ext"
  
)

cor_matrix<-round(cor(df_sample[,interested_variables],use = 'pairwise.complete.obs'),2)
corrplot(cor_matrix,type='upper',order='hclust',tl.cex=0.6)

df_cluster<-t(df_sample[complete.cases(df[,interested_variables]),interested_variables])
df_cluster<-scale(df_cluster)
dist_matrix<-dist(df_cluster, method='euclidean')
dist_mat <- as.matrix(dist_matrix)


melted_dist <- reshape2::melt(dist_mat)
ggplot(melted_dist, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "red", high = "white") +
  theme_minimal() +
  labs(title = "Euclidean Distance Matrix", x = "", y = "")

### difference SDQ CBCL/YSR
cor.test(df_sample$YSR_INT,df_sample$SDQ_self_int)
cor.test(df_sample$CBCL_INT,df_sample$SDQ_parent_int)
cor.test(df_sample$YSR_EXT,df_sample$SDQ_self_ext)
cor.test(df_sample$CBCL_EXT,df_sample$SDQ_parent_ext)

#concordance
cor.test(df_sample$YSR_INT,df_sample$CBCL_INT)
cor.test(df_sample$YSR_EXT,df_sample$CBCL_EXT)
cor.test(df_sample$SDQ_parent_int,df_sample$SDQ_self_int)
cor.test(df_sample$SDQ_parent_ext,df_sample$SDQ_self_ext)


```

### correlation table

- saliva cortisol is negatively associated with baseline pupil sizes
- hair cortisol is negatively assoicated with self- and parent-reported internalizing 
- SEPR average is negatively associated with self- and parent-reported externalizing

- BPS average is negatively associated with saliva cortisol 


```{r correlation_table}

interested_variables<-c(
  "saliva_Average.1",
  "saliva_Average.2",
  'saliva_change',
  "hair_cortisol_pg_mg", 
  "BPS_change",
  'BPS_average',
  "SEPR_average",
  "SEPR_Oddball_change",
  "INT_parent",
  "EXT_parent",
  "CBCL_T_GES",
  "INT_self",
  "EXT_self",
  "YSR_T_GES"
)

cor_matrix<-round(cor(df[,interested_variables],use = 'pairwise.complete.obs'),2)
corrplot(cor_matrix,type='upper',order='hclust',tl.cex=0.6)

```

### visualization

- higher psychopathology seems to be associated with attenuated saliva cortisol drop-off during the experiment

```{r visualization}

grid.arrange(
ggplot(df,aes(x=CBCL_T_GES,y=saliva_change))+geom_point()+geom_smooth(color='grey',method='lm'),
ggplot(df,aes(x=YSR_T_GES,y=saliva_change))+geom_point()+geom_smooth(color='grey',method='lm'),
ggplot(df,aes(x=YSR_T_INT,y=saliva_change))+geom_point()+geom_smooth(color='grey',method='lm'),
ggplot(df,aes(x=YSR_T_EXT,y=saliva_change))+geom_point()+geom_smooth(color='grey',method='lm'),
ggplot(df,aes(x=SDQ_total_parent,y=saliva_change))+geom_point()+geom_smooth(color='grey',method='lm'),
ggplot(df,aes(x=SDQ_total_self,y=saliva_change))+geom_point()+geom_smooth(color='grey',method='lm')
)

```

### covariate analysis

```{r covariate analysis}


#scale numeric variables
df_scaled<-as.data.frame(lapply(df_sample, function(x) {if (is.numeric(x)) scale(x) else x}))

###effects on psychopathology
lm<-lm(YSR_GES~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on self-reported psychopathology (YSR).",
           figure_name = 'covariate_effects_on_selfreport_psychopathology_YSR')

lm<-lm(SDQ_total_self~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on self-reported psychopathology (SDQ-s).",
           figure_name = 'covariate_effects_on_selfreport_psychopathology_SDQs')


lm<-lm(CBCL_GES~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm,
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on caregiver-reported psychopathology (CBCL).",
           figure_name = 'covariate_effects_on_caregiverreport_psychopathology_CBCL')

lm<-lm(SDQ_total_parent~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm,
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on caregiver-reported psychopathology (SDQ-p).",
           figure_name = 'covariate_effects_on_caregiverreport_psychopathology_SDQp')

##effects on pupillary response
lm<-lm(BPS_average~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on baseline pupil size (BPS).",
           figure_name = 'covariate_effects_on_BPS')

lm<-lm(SEPR_average~gender+age+iq+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = F)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','autism diagnosis','Residuals'),
           caption_label = "Covariate Effects on stimulus-evoked pupillary response (SEPR).",
           figure_name = 'covariate_effects_on_SEPR')

##effects on bio samples
lm<-lm(saliva_change_log~gender+age+iq+BMI+autism_boolean+saliva_CV.2+saliva_Time.1,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','BMI','autism diagnosis','measurement variation index','time of day','Residuals'),
           caption_label = "Covariate Effects on saliva cortisol concentration (SCC) change.",
           figure_name = 'covariate_effects_on_SCC')

lm<-lm(hair_cortisol_pg_mg_log~gender+age+iq+BMI+autism_boolean+hair_mass_mg,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','gender','age','IQ','BMI','autism diagnosis','hair mass','Residuals'),
           caption_label = "Covariate Effects on hair cortisol concentration (HCC).",
           figure_name = 'covariate_effects_on_HCC')


```

### bivariate association of pupillometry and cortisol concentration 

investigating the magnitude and reactivity of pupillary measures

```{r pupillometry with bio samples}

#scale numeric variables
df_scaled<-as.data.frame(lapply(df_sample, function(x) {if (is.numeric(x)) scale(x) else x}))

lm<-lm(SEPR_average~BPS_average+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','BPS','age','gender','autism diagnosis','Residuals'),
           caption_label = "Bivariate association of baseline pupil size (BPS) and stimulus-evoked pupillary response (SEPR).",
           figure_name = 'Bivariate_assoication_BPS_SEPR')

lm<-lm(hair_cortisol_pg_mg_log~saliva_change_log+
         age+gender+autism_boolean+hair_mass_mg,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','SCC change','age','gender','autism diagnosis','hair mass','Residuals'),
           caption_label = "Bivariate association of saliva cortisol concentration (SCC) change and hair cortisol concentration (HCC).",
           figure_name = 'Bivariate_assoication_SCCchange_HCC')


lm<-lm(hair_cortisol_pg_mg_log~saliva_Average.1_log+saliva_Average.2_log+
         age+gender+hair_mass_mg+autism_boolean,data=df_scaled)
analyze_lm(lm)


lm<-lm(BPS_average~hair_cortisol_pg_mg_log+saliva_change_log+
         age+gender+hair_mass_mg+autism_boolean,data=df_scaled)
analyze_lm(lm)

lm<-lm(SEPR_average~hair_cortisol_pg_mg_log+saliva_change_log+
         age+gender+hair_mass_mg+autism_boolean,data=df_scaled)
analyze_lm(lm)

```

### effect of psychopathology on cortisol

- higher self-reported psychopathology is associated with attenuated hair cortisol (hypocortisolism)
- even when controlling for autism symptoms in the autism-enriched sample
- can be attributed to internalizing versus externalizing symptoms
- also after controlling for relevant covariates

```{r psych cortisol}

#scale numeric variables
df_scaled<-as.data.frame(lapply(df_sample, function(x) {if (is.numeric(x)) scale(x) else x}))

lm<-lm(hair_cortisol_pg_mg_log~YSR_GES+hair_mass_mg+age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','YSR total','hair mass','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (YSR total) on hair cortisol concentration (HCC).",
           figure_name = 'YSRtotal_effects_on_HCC')

lm<-lm(hair_cortisol_pg_mg_log~SDQ_total_self+hair_mass_mg+age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','SDQ total','hair mass','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (SDQ total) on hair cortisol concentration (HCC).",
           figure_name = 'SDQtotal_effects_on_HCC')

lm<-lm(hair_cortisol_pg_mg_log~YSR_INT+YSR_EXT+hair_mass_mg+age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','YSR INT','YSR EXT','hair mass','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology subscales (YSR Internalizing, YSR Externalizing) on hair cortisol concentration (HCC).",
           figure_name = 'YSRsubscales_effects_on_HCC')

lm<-lm(hair_cortisol_pg_mg_log~SDQ_self_int+SDQ_self_ext+age+gender+hair_mass_mg+autism_boolean,data=df_scaled)
analyze_lm(lm,log_backtransform = T)
save_model(lm, 
           var_names = c('Intercept','SDQ INT','SDQ EXT','hair mass','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology subscales (SDQ Internalizing, SDQ Externalizing) on hair cortisol concentration (HCC).",
           figure_name = 'SDQsubscales_effects_on_HCC')


grid.arrange(
ggplot(df,aes(x=YSR_T_INT,y=hair_cortisol_pg_mg_log,alpha=YSR_T_GES))+geom_point()+geom_smooth(method='lm')+
  labs(title='hair cortisol and internalizing')+xlab('YSR, t-score')+ylab('hair cortisol (pg/mg, log)')+ theme(legend.position="none"),
ggplot(df,aes(x=YSR_T_EXT,y=hair_cortisol_pg_mg_log,alpha=YSR_T_GES))+geom_point()+geom_smooth(method='lm',col='red')+
    labs(title='hair cortisol and externalizing')+xlab('YSR, t-score')+ylab('hair cortisol (pg/mg, log)')+ theme(legend.position="none"),
ncol=2)


# SCC change
lm<-lm(saliva_change_log~YSR_T_GES+age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
lm<-lm(saliva_change_log~SDQ_total_self+age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)


```

### effect of psychopathology on pupillary measures

- high self-reported externalizing is associated with attenuated SEPR

```{r}

lm<-lm(SEPR_average~YSR_GES+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','YSR total','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (YSR total) on stimulus-evoked pupillary response (SEPR).",
           figure_name = 'YSRtotal_effects_on_SEPR')

lm<-lm(SEPR_average~SDQ_total_self+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','SDQ-s total','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (SDQ-s total) on stimulus-evoked pupillary response (SEPR).",
           figure_name = 'SDQtotal_effects_on_SEPR')

lm<-lm(SEPR_average~YSR_INT*YSR_EXT+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','YSR INT','YSR EXT','age','gender',
                         'autism diagnosis','YSR INT x YSR EXT','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (YSR Internalizing, YSR Externalizing) on stimulus-evoked pupillary response (SEPR).",
           figure_name = 'YSRsubscales_effects_on_SEPR')

lm<-lm(SEPR_average~SDQ_self_int*SDQ_self_ext+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','SDQ-s INT','SDQ-s EXT','age','gender',
                         'autism diagnosis','SDQ-s INT x SDQ-s EXT','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (SDQ-s Internalizing, SDQ-s Externalizing) on stimulus-evoked pupillary response (SEPR).",
           figure_name = 'SDQsubscales_effects_on_SEPR')



lm<-lm(BPS_average~YSR_GES+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','YSR total','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (YSR Total) on baseline pupil size (BPS).",
           figure_name = 'YSRtotal_effects_on_BPS')

lm<-lm(BPS_average~SDQ_total_self+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','SDQ-s total','age','gender',
                         'autism diagnosis','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (SDQ-s Total) on baseline pupil size (BPS).",
           figure_name = 'SDQtotal_effects_on_BPS')

lm<-lm(BPS_average~YSR_INT*YSR_EXT+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','YSR INT','YSR EXT','age','gender',
                         'autism diagnosis','YSR INT x YSR EXT','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (YSR Internalizing, YSR Externalizing) on baseline pupil size (BPS).",
           figure_name = 'YSRsubscales_effects_on_BPS')

lm<-lm(BPS_average~SDQ_self_int*SDQ_self_ext+
         age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
save_model(lm, 
           var_names = c('Intercept','SDQ-s INT','SDQ-s EXT','age','gender',
                         'autism diagnosis','SDQ-s INT x SDQ-s EXT','Residuals'),
           caption_label = "Linear model - effect of self-reported psychopathology (SDQ-s Internalizing, SDQ-s Externalizing) on baseline pupil size (BPS).",
           figure_name = 'SDQsubscales_effects_on_BPS')


```

### cortisol and pupillary measures explaining self-reported internalizing

- lower hair cortisol associated with higher internalizing psychopathology
- higher BPS is associated with higher internalizing
- the interaction suggests that this might be specifically relevant in groups of elevated symptoms
- pupillary measures might be a quantitative predictor of internalizing psychopathology in at-risk children

```{r measures on internalizing}

lm<-lm(YSR_INT~BPS_average*SEPR_average*hair_cortisol_pg_mg*saliva_change+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC change',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC change','SEPR x SCC change',
                         'HCC x SCC change','BPS x SEPR x HCC','BPS x SEPR x SCC change',
                         'BPS x HCC x SCC change','SEPR x HCC x SCC change',
                         'BPS x SEPR x HCC x SCC change','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Internalizing (YSR Internalizing).",
           figure_name = 'combined_effects_on_YSRint')


### R² = 0.17
emtrends(lm, ~hair_cortisol_pg_mg+SEPR_average, var="BPS_average", 
         at = list(hair_cortisol_pg_mg = c(-1,1),SEPR_average = c(-1,1)))

  #residual normality
  res_fe<-residuals(lm)
  hist(res_fe)
  qqnorm(res_fe)
  
  #moderation by Externalizing
  lm<-lm(YSR_INT~BPS_average*SEPR_average*hair_cortisol_pg_mg*saliva_change+YSR_EXT+
         +age+gender+autism_boolean,data=df_scaled)
  analyze_lm(lm)

## SDQ Internalizing
### R² = 0.23
lm<-lm(SDQ_self_int~BPS_average*SEPR_average*hair_cortisol_pg_mg*saliva_change+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC change',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC change','SEPR x SCC change',
                         'HCC x SCC change','BPS x SEPR x HCC','BPS x SEPR x SCC change',
                         'BPS x HCC x SCC change','SEPR x HCC x SCC change',
                         'BPS x SEPR x HCC x SCC change','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Internalizing (SDQ-s Internalizing).",
           figure_name = 'combined_effects_on_SDQint')


emtrends(lm, ~hair_cortisol_pg_mg+SEPR_average, var="BPS_average", at = list(hair_cortisol_pg_mg = c(-1,0,1),SEPR_average = c(-1,0,1)))


  #moderation by Externalizing
  lm<-lm(SDQ_self_int~BPS_average*SEPR_average*hair_cortisol_pg_mg*saliva_change+SDQ_self_ext+
           +age+gender+autism_boolean,data=df_scaled)
  analyze_lm(lm)


tiff(file=paste0(project_path,"/output/figures/figure_pupillometry_HCC_Internalizing.tiff"), 
     width=9, height=8, units="in", res=600, compression='lzw') 

#YSR scatterplot
df_scaled$SEPR_average_split<-ifelse(df_scaled$SEPR_average>=0,'above average SEPR','below average SEPR')
df_scaled$HCC_average_split<-ifelse(df_scaled$hair_cortisol_pg_mg>=0,'above average HCC','below average HCC')
g1<-ggplot(df_scaled[!is.na(df_scaled$SEPR_average_split) & !is.na(df_scaled$HCC_average_split),],
       aes(x=BPS_average,y=YSR_INT,color=HCC_average_split))+
  geom_point(size=2)+
  geom_smooth(method='lm',color='darkgrey')+
  facet_wrap(~SEPR_average_split+HCC_average_split)+
  xlab('baseline pupil size (BPS, in z)')+
  ylab('YSR Internalizing (z)')+
  theme(legend.position = 'none')
  
#YSR by groups
df_scaled$BPS_average_split<-ifelse(df_scaled$BPS_average>=0,'above average BPS','below average BPS')
df_plot<-plyr::rbind.fill(by(df_scaled$YSR_T_INT,interaction(df_scaled$HCC_average_split,df_scaled$BPS_average_split),psych::describe))
df_plot$HCC<-c('above average','below average','above average','below average')
df_plot$pupilsize<-c('above average','above average','below average','below average')

g2<-ggplot(df_plot,aes(x=pupilsize,y=mean,fill=HCC))+
  geom_bar(position=position_dodge(), stat="identity", colour='black') +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.2,position=position_dodge(.9))+
  geom_text(aes(label=paste('N=',n)), vjust=2, hjust=0.5,position = position_dodge(width = .9))+
  xlab('BPS (median split)')+ylab('YSR Internalizing (z)')

#SDQ by scatter
df_scaled$SEPR_average_split<-ifelse(df_scaled$SEPR_average>=0,'above average SEPR','below average SEPR')
df_scaled$HCC_average_split<-ifelse(df_scaled$hair_cortisol_pg_mg>=0,'above average HCC','below average HCC')
g3<-ggplot(df_scaled[!is.na(df_scaled$SEPR_average_split) & !is.na(df_scaled$HCC_average_split),],
       aes(x=BPS_average,y=SDQ_self_int,color=HCC_average_split))+
  geom_point(size=2)+
  geom_smooth(method='lm',color='darkgrey')+
  facet_wrap(~SEPR_average_split+HCC_average_split)+
  xlab('baseline pupil size (BPS, in z)')+
  ylab('SDQ Internalizing (z)')+
  theme(legend.position = 'none')

#SDQ by groups
df_scaled$BPS_average_split<-ifelse(df_scaled$BPS_average>=0,'above average BPS','below average BPS')
df_plot<-plyr::rbind.fill(by(df_scaled$SDQ_self_int,interaction(df_scaled$HCC_average_split,df_scaled$BPS_average_split),psych::describe))
df_plot$HCC<-c('above average','below average','above average','below average')
df_plot$pupilsize<-c('above average','above average','below average','below average')

g4<-ggplot(df_plot,aes(x=pupilsize,y=mean,fill=HCC))+
  geom_bar(position=position_dodge(), stat="identity", colour='black') +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.2,position=position_dodge(.9))+
  geom_text(aes(label=paste('N=',n)), vjust=2, hjust=0.5,position = position_dodge(width = .9))+
  xlab('BPS (median split)')+ylab('SDQ Internalizing (z)')

grid.arrange(g1,g2,g3,g4,ncol=2,nrow=2,widths=c(4,3))

dev.off()
  

```

### cortisol and pupillary measures explaining self-reported externalizing

- specific to SDQ externalizing
- higher hair cortisol also associated with attenuated self-reported externalizing
- however --> mediated by pupillary responses interaction
- at above average BPS levels - lower SEPR predict elevated externalizing

- in parent reports - lower SEPR is associated with increased externalizing

```{r measures on externalizing}

#YSR Ext
lm<-lm(YSR_EXT~BPS_average*SEPR_average*hair_cortisol_pg_mg*saliva_change+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC change',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC change','SEPR x SCC change',
                         'HCC x SCC change','BPS x SEPR x HCC','BPS x SEPR x SCC change',
                         'BPS x HCC x SCC change','SEPR x HCC x SCC change',
                         'BPS x SEPR x HCC x SCC change','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Externalizing (YSR).",
           figure_name = 'combined_effects_on_YSRext')



### R² = 0.07
emtrends(lm, ~SEPR_average+hair_cortisol_pg_mg, var="saliva_change", at = list(SEPR_average = c(-1,1),hair_cortisol_pg_mg = c(-1,1)))

#SDQ Ext
lm<-lm(SDQ_self_ext~BPS_average*SEPR_average*hair_cortisol_pg_mg*saliva_change+
         +age+gender+autism_boolean,data=df_scaled)
analyze_lm(lm)
### R² = 0.07

save_model(lm, 
           var_names = c('Intercept','BPS','SEPR','HCC','SCC change',
                         'age','gender','autism diagnosis',
                         'BPS x SEPR','BPS x HCC','SEPR x HCC','BPS x SCC change','SEPR x SCC change',
                         'HCC x SCC change','BPS x SEPR x HCC','BPS x SEPR x SCC change',
                         'BPS x HCC x SCC change','SEPR x HCC x SCC change',
                         'BPS x SEPR x HCC x SCC change','Residuals'),
           caption_label = "Linear model - combined effects of pupillometry (BPS, SEPR) and cortisol concentration (HCC, SCC change) on self-reported Externalizing (SDQ-s).",
           figure_name = 'combined_effects_on_SDQext')

# df_scaled$SEPR_average_split<-ifelse(df_scaled$SEPR_average>=0,'above average SEPR','below average SEPR')
# df_scaled$HCC_average_split<-ifelse(df_scaled$hair_cortisol_pg_mg>=0,'above average HCC','below average HCC')
# ggplot(df_scaled[!is.na(df_scaled$SEPR_average_split) & !is.na(df_scaled$HCC_average_split),],
#        aes(x=saliva_change,y=YSR_EXT,color=HCC_average_split))+
#   geom_point(size=2)+
#   geom_smooth(method='lm',color='darkgrey')+
#   facet_wrap(~SEPR_average_split+HCC_average_split)+
#   xlab('saliva_change (HCC change, in z)')+
#   ylab('SDQ Externalizing (z)')+
#   theme(legend.position = 'none')



# lm1<-lm(SDQ_self_ext~hair_cortisol_pg_mg+
#          age+autism_boolean,data=df_scaled)
# analyze_lm(lm1)
# summary(lm1)
# 
# lm1<-lm(SDQ_self_ext~BPS_average*SEPR_average+
#          age+autism_boolean,data=df_scaled)
# analyze_lm(lm1)
# summary(lm1)
# ### R² = 0.09
# emtrends(lm1, ~BPS_average, var="SEPR_average", at = list(BPS_average = c(-1,0,1)))
# 
# 
# df_scaled$BPS_average_split<-ifelse(df_scaled$BPS_average>=0,'high','low')
# ggplot(df_scaled,aes(x=SEPR_average,y=SDQ_self_ext,group=BPS_average_split,fill=BPS_average_split,color=BPS_average_split))+
#   geom_smooth(method='lm')+geom_point()+
#   xlab('Pupillary Response (z)')+ylab('self-reported externalizing (z)')


```

### classification

- above change classification based on pupillometry and cortisol measures


```{r classification, warnings=F}
set.seed(333)

split_ratio<-0.1


df_sample$YSR_INT_CUTOFF<-ifelse(df_sample$YSR_T_INT>=70,T,F)
df_sample$YSR_EXT_CUTOFF<-ifelse(df_sample$YSR_T_EXT>=70,T,F)

# Ensure the target variable is a factor
df_sample$YSR_INT_CUTOFF <- as.factor(df_sample$YSR_INT_CUTOFF)
df_sample$YSR_EXT_CUTOFF <- as.factor(df_sample$YSR_EXT_CUTOFF)


table(df_sample$YSR_INT_CUTOFF)
table(df_sample$YSR_EXT_CUTOFF)

#complete data
df_sample_complete<-df_sample[complete.cases(df_sample$BPS_average) & 
                  complete.cases(df_sample$SEPR_average) & 
                  complete.cases(df_sample$hair_cortisol_pg_mg) & 
                  complete.cases(df_sample$saliva_change) & complete.cases(df_sample$YSR_INT_CUTOFF),]


#bootstrapping function - cortisol
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_INT_CUTOFF ~ hair_cortisol_pg_mg*saliva_change, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}

# bootstrapping
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("cortisol AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


#bootstrapping function - pupillometry
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_INT_CUTOFF ~ BPS_average*SEPR_average, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}


# Combine training data and testData for bootstrapping
# (You can also bootstrap on testData if preferred)
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("pupillometry AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


# Bootstrapping function to compute AUC
# boot_auc <- function(data, indices) {
#   d <- data[indices, ]
#   model <- glm(YSR_INT_CUTOFF ~ BPS_average*SEPR_average*hair_cortisol_pg_mg, data = d, family = binomial)
#   probs <- predict(model, newdata = testData, type = "response")
#   roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
#   return(auc(roc_obj))
# }

#bootstrapping function - pupillometry + hair cortisol
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
  model <- glm(YSR_INT_CUTOFF ~ BPS_average*SEPR_average*hair_cortisol_pg_mg, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}

# Combine training data and testData for bootstrapping
# (You can also bootstrap on testData if preferred)
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("pupillometry + hair cortisol AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)


# Bootstrapping function to compute AUC
boot_auc <- function(data, indices) {
  trainIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  testIndex <- createDataPartition(data$YSR_INT_CUTOFF, p = split_ratio, list = FALSE)
  trainData <- df_sample_complete[-trainIndex, ]
  testData <- df_sample_complete[-testIndex, ]
   model <- glm(YSR_INT_CUTOFF ~ BPS_average*SEPR_average*hair_cortisol_pg_mg, data = df_sample_complete, family = binomial)
  probs <- predict(model, newdata = testData, type = "response")
  roc_obj <- roc(testData$YSR_INT_CUTOFF, probs)
  return(auc(roc_obj))
}

# Combine training data and testData for bootstrapping
# (You can also bootstrap on testData if preferred)
suppressMessages({boot_results <- boot(data = df_sample_complete, statistic = boot_auc, R = 1000)
})

# Confidence interval for AUC
ci <- boot.ci(boot_results, type = "perc")
print(paste("pupillometry + cortisol AUC:", round(mean(boot_results$t,na.rm=T),3)))
print(ci)




```


### trial-based analysis
```{r}

ggplot(df_trial_visualoddball,aes(x=trial_number,y=pd_low,group=id))+geom_smooth(method='lm')

require(lme4)
require(lmerTest)

#estimate slope of BPS change within visual oddball task
lmm<-lmer(pd_low~trial_number*id+(1|id),df_trial_visualoddball)
anova(lmm)

#extract data
df_BPSslope<-data.frame(emtrends(lmm,~id,var='trial_number'))
df_BPSslope<-df_BPSslope[,1:3]
names(df_BPSslope)<-c('id','BPS_slope','BPS_slope_SD')
hist(scale(df_BPSslope$BPS_slope))
summary(scale(df_BPSslope$BPS_slope))

#merge with df
df$merger_id<-as.numeric(substr(df$SEGA_ID,6,8))
df<-merge(df,df_BPSslope,by.x='merger_id',by.y='id',all.x=T)


lm<-lm(scale(BPS_slope)~YSR_T_INT*YSR_T_EXT+
         age+autism_boolean,data=df)
summary(lm)


```

### classification based on condition data

```{r}

names(df_sample)

df_class<-df_sample[,57:88]

#impute missings with rowwise mean
df_class_imputed <- df_class
for (j in seq_along(df_class_imputed)) {
  na_idx <- is.na(df_class_imputed[[j]])
  if (any(na_idx)) {
    df_class_imputed[na_idx, j] <- median(df_class_imputed[[j]], na.rm = TRUE)
  }
}

# Euclidean distance
d <- dist(df_class_imputed, method = "euclidean")

# Hierarchical clustering with Ward's method (recommended for many numeric datasets)
hc <- hclust(d, method = "ward.D2")

# Base dendrogram
plot(hc, cex = 0.6, hang = -1, main = "Hierarchical Clustering Dendrogram (Ward.D2)")

#assumed clusters based on dendrogram
clusters <- cutree(hc, k = 3)

# Append cluster labels to your original (complete) data
result <- data.frame(df_sample, cluster = factor(clusters))

with(result,by(YSR_T_INT,cluster,psych::describe))
with(result,by(YSR_T_EXT,cluster,psych::describe))
with(result,by(YSR_T_GES,cluster,psych::describe))

```


